---
title: "02: Exploratory Data Analysis for Singapore CPI (2020 to 2025)"
author: "Yiqiong PAN"
date: "24 Feb, 2026"
date-modified: last-modified

execute:
  echo: true    # display the code
  eval: true    # run the code
  message: false
  warning: false
  freeze: true  # do not re-render if nothing edited

format: 
  html:
    toc: true
    code-link: true

number-sections: true
number-depth: 9

editor: visual
---

## Introduction

The Consumer Price Index (CPI) is a widely used economic indicator that measures inflation by tracking changes in the price level of a basket of goods and services consumed by households. CPI provides an important reference for economic policy, wage adjustments, and cost of living assessments. In Singapore, CPI statistics are compiled and published by the Singapore Department of Statistics and are structured into 10 expenditure categories such as food, housing, transport, healthcare, and recreation.

The CPI dataset naturally forms a **hierarchical** structure, where detailed consumption items aggregate into broader categories and ultimately contribute to the overall inflation index. For example, individual items such as rice, meat, and vegetables belong to lower level sub categories, which aggregate into food related groups and eventually into the overall CPI. 

This study examines monthly Singapore CPI data from 2020 to 2025, covering multiple CPI categories from the overall index to detailed expenditure components. The period from **2020 to 2025** represents an especially meaningful timeframe for inflation analysis. During this period, the global economy experienced significant disruptions due to the COVID 19 pandemic, supply chain constraints, and subsequent economic recovery. These factors contributed to changes in consumption patterns and price volatility across different sectors. Analysing CPI during this period provides insights into how inflation dynamics evolved under these unusual economic conditions. By combining exploratory visualisation and time series forecasting, the study aims to understand the behaviour of inflation across different CPI components and generate forecasts for future price movements.


## Motivation

While headline inflation measures such as the overall CPI provide a broad indication of price changes, they often conceal important differences across consumption categories. Certain components, such as food or energy related items, may experience stronger volatility or seasonal patterns compared with more stable categories such as healthcare or education. Analysing CPI at a more detailed level therefore allows a clearer understanding of which sectors contribute most to inflation changes.

Visualisation plays an important role in exploring CPI data. Time series plots, seasonal visualisations, and autocorrelation analysis help reveal underlying patterns such as trends, seasonal behaviour, and persistence in price movements. These exploratory insights are valuable for understanding the structure of CPI data and informing the selection of appropriate forecasting models.

Another challenge arises from the fact that CPI datasets often contain many related time series across different categories. In this study, multiple CPI components are analysed simultaneously, each representing a separate time series with its own price dynamics. Manually building and evaluating forecasting models for each series would be inefficient and difficult to maintain.

To address this challenge, the study adopts a nested time series forecasting approach using the Modeltime package. Nested forecasting allows each CPI series to be modelled independently within a unified workflow. Multiple forecasting models can be trained, evaluated, and compared automatically across all CPI categories, enabling efficient model selection and scalable forecasting.

This approach offers several advantages. First, it enables systematic modelling of a large number of CPI series while maintaining a consistent workflow. Second, it allows different models to compete for each series, ensuring that the most suitable model is selected based on performance. Finally, it provides a flexible framework that integrates statistical forecasting methods and machine learning techniques within a tidy data workflow.

By combining exploratory data visualisation with nested forecasting techniques, this study aims to provide a comprehensive analysis of Singapore’s CPI dynamics and produce reliable forecasts for its key components.

## Analytical Workflow

To analyse and forecast Singapore’s CPI data, a structured analytical workflow was implemented to ensure systematic exploration, modelling, and evaluation of the time series.

First, data preparation and transformation were conducted to organise the CPI dataset into a tidy time series structure suitable for analysis. Monthly CPI data from 2020 to 2025 was reshaped into a long format so that each CPI category represents an individual time series. This structure enables consistent analysis across multiple CPI components.

Second, exploratory data analysis (EDA) was performed to understand the statistical properties of the CPI series. Time series visualisations were used to examine long term trends and variability across categories. In addition, time series decomposition was applied to separate each series into its trend, seasonal, and irregular components, allowing clearer interpretation of underlying inflation patterns. Autocorrelation and partial autocorrelation analysis were also conducted to assess temporal dependence and provide insights into appropriate model structures.

Third, a nested forecasting workflow was implemented using the Modeltime package. In this approach, each CPI category is treated as an individual time series within a nested data structure. Multiple forecasting models are trained and evaluated automatically across all series, enabling efficient model comparison and scalable forecasting.


## Getting Started

### Installing and Loading the Required Libraries

There are <mark>**XX**</mark> packages that will be used in this exercise.


| Package | Description |
|------------------------------------|------------------------------------|
| [tidyverse](https://www.tidyverse.org/) | A collection of R packages for data importing, cleaning, transformation, and visualisation (e.g., [readr](https://readr.tidyverse.org/), dplyr, tidyr, [ggplot2](https://ggplot2.tidyverse.org/)). |
| [knitr](https://cran.r-project.org/web/packages/knitr/index.html/) | Provides a general-purpose tool for dynamic report generation in R using Literate Programming techniques. |
| [DT](https://rstudio.github.io/DT/) | Displays data frames as interactive HTML tables with sorting, filtering, and pagination. |
| [patchwork](https://cran.r-project.org/web/packages/patchwork/index.html) | Simpily combine separate ggplots into the same graphic. |
| [scales](https://cran.r-project.org/web/packages/scales/index.html) | Provides graphical scales map data to aesthetics. |
| [magick](https://cran.r-project.org/web/packages/magick/index.html) | Offer advanced graphics and image-Processing in R. |
| [ggthemes](https://cran.r-project.org/web/packages/ggthemes/index.html) | Provides 'ggplot2' themes and scales that replicate some of the famous looks. |
| [readxl](https://cran.r-project.org/web/packages/readxl/readxl.pdf) | Imports excel files into R. |
| [gganimate](https://gganimate.com/) | The ggplot2 extension used to create animated statistical graphs. |
| [gifski](https://cran.r-project.org/web/packages/gifski/index.html) | A rendering engine used by gganimate to turn animation frames into high quality GIFs. It handles the final export, not the animation logic itself. |
| [plotly](https://plotly.com/r/) | Creates interactive and publication-ready graphs. |
| [gapminder](https://cran.r-project.org/web/packages/gapminder/index.html) | A dataset and visualisation concept popularised by Hans Rosling. In ggplot and gganimate, Gapminder data is often used as an example to demonstrate animated plots over time, such as countries moving across income and life expectancy.Used here solely to apply its country color scheme. |
| [tidymodels](https://www.tidymodels.org/) | A unified modelling framework for machine learning in R |
| [timetk](https://github.com/business-science/timetk) | A cutting-edge toolkit for time series preparation and exploration. |
| [modeltime](https://business-science.github.io/modeltime/index.html) | A new forecasting framework that simplifies the full workflow of building time series models (such as ARIMA, exponential smoothing, and machine learning models) |
| [lubridate](https://cran.r-project.org/web/packages/lubridate/index.html) | A Tidyverse package that handles date-time objects |

The code below installs and loads the required packages into R environment.

```{r}
pacman::p_load(tidyverse, DT, readxl,
             knitr,patchwork, scales,magick, ggthemes,gifski, gapminder,
               plotly, gganimate,tidymodels,modeltime, timetk, lubridate, stringr, htmltools, viridis, rlang)
```

### Importing the Data

The dataset was obtained from the [CEIC Database](https://www.ceicdata.com/en) via the SMU Library. The Excel file contains monthly Consumer Price Index (CPI) data with base year 2024, spanning from January 1961 to January 2026,last updated on 23/02/2026.

The dataset is imported into R using the [`read_excel()`](https://readxl.tidyverse.org/reference/read_excel.html) function from the **readxl** package and stored as a tibble named `df`. The first few rows of the dataset are then displayed using `kable()` for inspection.

```{r}
df <- read_excel("data/Series.xlsx")
```

```{r}
kable(head(df, 10), "html")
```

When downloading the dataset, only the Data Column: **Series Name** option was selected in the output settings. As a result, the metadata rows were not included in the downloaded file. If the metadata rows are present, they could be removed using the `slice()` function, for example `slice(-1:28)`, to exclude the first 28 rows from the dataset.

The dataset contains **208 columns and 781 rows**. The column names are not well structured and contain inconsistent formatting and lengthy labels. Therefore, further cleaning and standardisation are required before the dataset can be used for analysis.

## Data Wrangling

### Renaming Date Column and filtering the Study Period

The code below uses the `rename()` function to replace the unnecessarily long column name with `date`. It also applies the **`filter()` function from `dplyr`** to keep observations between **2020 and 2025** for exploratory data analysis (EDA). This ensures consistency with the project scope, which focuses on **demonstration and forecasting**. Although **data for January 2026 has already been released**, it is excluded at this stage to maintain a consistent analysis period.

```{r}
df <- df %>%
  rename(date = 1) %>%
  mutate(date = as_date(date)) %>%
  filter(date >= "2020-01-01" & date <= "2025-12-31")
```

### Reshaping the Data

Since this project focuses on **time series analysis**, the original CPI dataset in **wide format** was transformed into a **long format** to make it suitable for time series modelling and analysis.

The code below uses the `pivot_longer()` function to transform the dataset from wide format to long format.

```{r}
df_long <- df %>%
  pivot_longer(
    cols = -date,
    names_to = "series",
    values_to = "cpi"
  )
```

### Labelling CPI Levels

The code block below saves and displays all CPI items in order to determine the desired table structure. Here we display the first 20 CPI items.

```{r}
cpi_items <- df %>%
  names() %>%
  setdiff("date")

length(cpi_items)
cpi_items[1:20]
```

To ensure the structure aligns with the official CPI classification, the [**Department of Statistics (DOS) CPI documentation**](https://www.singstat.gov.sg/-/media/files/news/cpijan26.ashx) was used as a reference. All published CPI items and their corresponding hierarchical levels were extracted according to the official structure.

The dataset preserves the full level of granularity defined by the official CPI classification. As a result, some combinations contain **missing values**, which is expected because not every subcategory is reported for all periods.

The code below reveals that we have 2064 missing values in the dataset.

```{r}
sum(is.na(df))
```

After the transformation and extraction process, the CPI items in the dataset now match the official DOS CPI classification structure.

After the transformation, the CPI hierarchy is identified by counting the number of **colon (`:`) occurrences** in the item names by using `str_count()` function from **stringr** package.

A new column called `level` is created to represent the hierarchical structure of the CPI categories. For example:

-   `Consumer Price Index (CPI)` is assigned **level 0**

-   Major groups such as **Food** and **Health** are assigned **level 1**

-   Their subcategories are assigned **level 2**

```{r}
df_long <- df_long %>%
 mutate(
    level = str_count(series, ":")) %>%
  filter(level <= 2)
```

However, several exceptions must be handled after the initial level assignment. While the number of colon occurrences provides a useful starting point for identifying CPI hierarchy levels, it does not fully align with the official DOS classification.

In particular, categories under **Food** require manual adjustment: **Food** is classified as level 1, while `CPI: Food Excl Food & Beverage Serving Services (FFBSS)` and `CPI: Food & Beverage Serving Services (FBSS)` should be reclassified as level 2, with their nested subgroups assigned to level 3.

In addition, categories beginning with **All Items Less** should be treated as level 0. These refinements ensure that the final hierarchy is consistent with the official CPI classification structure.

The code chunk below uses `str_detect()` from the **stringr** package to identify specific CPI categories using regular expressions and revise the assigned CPI levels to match the official published classification structure.

```{r}
df_long <- df_long %>%
  mutate(
    level = case_when(
      str_detect(series, "^CPI: All Items Less") ~ 0,

      series %in% c(
        "CPI: Food Excl Food & Beverage Serving Services (FFBSS)",
        "CPI: Food & Beverage Serving Services (FBSS)"
      ) ~ 2,

      str_detect(series, "^CPI: FFBSS") &
        str_count(series, ":") >= 2 ~ str_count(series, ":") + 1,

      str_detect(series, "^CPI: FBSS") &
        str_count(series, ":") >= 2 ~ str_count(series, ":") + 1,

      TRUE ~ level
    )
  )
```

### Standardising CPI names

We use `str_remove()` to remove the **“CPI:” prefix** from the series names. This helps clean up the labels so the table does not repeatedly display “CPI:” in every series name, making the series easier to read and rename the file to `cpi`.

```{r}
cpi <- df_long %>%
  mutate(series = str_remove(series, "^CPI: "))
```

### Sanity Check

Here, we take a quick look at the structure of the CPI dataset using the `glimpse()` function from the **dplyr** package. We also perform a sanity check to confirm that there are no missing values in the dataset.

```{r}
glimpse(cpi)
```

```{r}
sum(is.na(cpi))
```

### Saving the File

Here we use the `write_rds()` function to save the cleaned CPI dataset. The file can later be reloaded into the system for further analysis.

```{r}
#| eval: false
write_rds(cpi, "data/cpi.rds")
write_csv(cpi, "data/cpi.csv")
```

```{r}
cpi<- read_rds("data/cpi.rds")
```

### Building an Interactive Table

Finally, we create a `DT` table to display the dataset in an interactive tabular format.

::: panel-tabset
#### Step 1: Creating Reference Table and Join

The first step is to build a reference table for the CPI hierarchy. Since the list of CPI series is finite, we first extract the **distinct series names** from the dataset to create a clean list without duplicated monthly records.

Next, we manually assign the hierarchy levels for each series, such as **division, group, and class**, to define the relationships in the CPI structure.

Finally, this annotated hierarchy table is **joined** back to the main dataset, so that each observation inherits the hierarchy columns.

```{r}
#| code-fold: true
#| code-summary: "SHOW"
series_meta <- cpi %>%
  distinct(series, level) %>%
  mutate(
    division = case_when(
      series == "Consumer Price Index (CPI)" ~ NA_character_,
      series == "Food" ~ "Food",
      series == "Clothing and Footwear (C&F)" ~ "Clothing and Footwear (C&F)",
      series == "Housing & Utilities" ~ "Housing & Utilities",
      series == "Household Durables & Services (HDS)" ~ "Household Durables & Services (HDS)",
      series == "Health" ~ "Health",
      series == "Transport" ~ "Transport",
      series == "Information & Communication (InfoComm)" ~ "Information & Communication (InfoComm)",
      series == "Recreation, Sport & Culture (RSC)" ~ "Recreation, Sport & Culture (RSC)",
      series == "Education" ~ "Education",
      series == "Miscellaneous Goods & Services (MG&S)" ~ "Miscellaneous Goods & Services (MG&S)",
      # groups/classes inherit division from their parent:
      grepl("^Food|FFBSS|FBSS", series) ~ "Food",
      grepl("^C&F:", series) ~ "Clothing and Footwear (C&F)",
      grepl("^Housing & Utilities", series) ~ "Housing & Utilities",
      grepl("^HDS:", series) ~ "Household Durables & Services (HDS)",
      grepl("^Health:", series) ~ "Health",
      grepl("^Transport:", series) ~ "Transport",
      grepl("^InfoComm:", series) ~ "Information & Communication (InfoComm)",
      grepl("^RSC:", series) ~ "Recreation, Sport & Culture (RSC)",
      grepl("^Education:", series) ~ "Education",
      grepl("^MG&S:", series) ~ "Miscellaneous Goods & Services (MG&S)",
      TRUE ~ NA_character_
    ),
    group = case_when(
      level == 2 ~ series,  # each level-2 row is a group
      level == 3 & grepl("^FFBSS:", series) ~
        "Food Excl Food & Beverage Serving Services (FFBSS)",
      level == 3 & grepl("^FBSS:", series) ~
        "Food & Beverage Serving Services (FBSS)",
      TRUE ~ NA_character_
    ),
    class = case_when(
      level == 3 ~ series,  # keep full label for classes
      TRUE ~ NA_character_
    )
  )

cpi_h <- cpi %>%
  left_join(series_meta, by = c("series", "level"))

```

#### Step 2: MOdifying the Table for Interactivity

Next we prepares the CPI dataset for display in an interactive table.

First, the data is sorted by `series` and `date`, then grouped by `series` so calculations are applied within each CPI series.

Within `mutate()`:

-   `As_of` keeps the original date for filtering, while `As_of_lab` creates a readable label such as *Jan 2024*.
-   `cpi_change` calculates the **month to month CPI change** using `lag()`.
-   `Trend` converts the change into symbols: ▲ increase, ▼ decrease, ◆ first observation or no change.
-   `level_label` converts numeric hierarchy levels into labels (Top Level, Major Category, Sub Category, Detailed).
-   `major_group` and `display_name` determine which hierarchy name (division, group, or class) should be shown.
-   `cpi_rounded` rounds CPI values to two decimals.

Finally, the dataset is **ungrouped** and selected columns are renamed to produce a clean table ready for interactive display.

```{r}
#| code-fold: true
#| code-summary: "SHOW"
#| 
cpi_h_display <- cpi_h %>%
  arrange(series, date) %>%
  group_by(series) %>%                     # for MoM change
  mutate(
    # raw Date for filtering
    As_of = date,
    As_of_lab = format(date, "%b %Y"),

    # month-on-month change (current - previous month)
    cpi_change = cpi - lag(cpi),

    # Trend symbol based on change
    Trend = case_when(
      is.na(cpi_change) ~ "◆",   # first observation
      cpi_change > 0  ~ "▲",
      cpi_change < 0 ~ "▼",
      TRUE ~ "◆"
    ),

    level_label = case_when(
      level == 0 ~ "Top Level",
      level == 1 ~ "Major Category",
      level == 2 ~ "Sub-Category",
      level == 3 ~ "Detailed",
      TRUE ~ paste("Level", level)
    ),
    major_group = case_when(
      level == 0 ~ NA_character_,
      TRUE       ~ division
    ),
    display_name = case_when(
      level == 0 ~ series,
      level == 1 ~ division,
      level == 2 ~ group,
      level == 3 ~ class,
      TRUE ~ series
    ),
    cpi_rounded = round(cpi, 2)
  ) %>%
  ungroup() %>%
  select(
    Level = level_label,
    Division= major_group,
    Group = group,
    Class = class,
    Series = display_name,
    `CPI Value`= cpi_rounded,
    `As of` = As_of,       # Date for filters
    `As of label` = As_of_lab, # pretty label
    Trend
  )

```

#### Step 3: Defining CSS

Because the table contains many columns and large fonts, it is difficult to display everything clearly on a single screen. To improve readability, we make the table more compact and enable responsive behaviour.

First, we add a small CSS block to **reduce** the font size and cell paddin\* in the DataTable. This allows more information to fit within the visible space without shrinking the entire browser view.

```{r}
#| code-fold: true
#| code-summary: "SHOW"
compact_dt_css <- tags$style(HTML("
  table.dataTable tbody td {
    padding: 4px 6px !important;
    font-size: 13px !important;
  }
  table.dataTable thead th {
    padding: 6px 6px !important;
    font-size: 13px !important;
  }
  div.dataTables_filter input {
    font-size: 13px;
  }
"))


```

#### Step 4: Defining JavaScript Render

First, a **JavaScript renderer (`trend_js`)** is defined. This function colours the trend symbols in the table: ▲ appears in green for an increase, ▼ appears in red for a decrease, and ◆ appears in grey for no change or the first observation. The symbols are also displayed in bold with a larger font size to make the trend visually clear.

```{r}
#| code-fold: true
#| code-summary: "SHOW"
# JS for colored trend symbols
trend_js <- JS(
  "function(data, type, row, meta) {",
  "  if(type === 'display') {",
  "    var color = data === '▲' ? 'green' : (data === '▼' ? 'red' : 'gray');",
  "    return '<span style=\"color:' + color + '; font-weight:bold; font-size:18px;\">' + data + '</span>';",
  "  }",
  "  return data;",
  "}"
)
```

#### Step 5: Putting Everything Together

Finally we builds a compact, interactive CPI table using the `DT` package and custom styling. It combines **JavaScript, HTML layout, and DataTable options** to make the CPI hierarchy easier to explore.

The table layout is created using `div()`. The `compact_dt_css` style is applied to reduce font size and padding so the table can fit more information on the screen. The container also limits the maximum width and centres the table for a cleaner layout.

A **title section** is added at the top to display the dashboard heading (“Singapore Consumer Price Index”) and the latest available data month, which is calculated dynamically from the dataset.

Below the title, a **summary bar** shows three key statistics: the total number of rows in the dataset, the number of distinct CPI divisions, and the latest update date. These values are calculated directly from `cpi_h_display`.

The **main interactive table** is created using `datatable()`. Only the relevant columns are displayed: hierarchy levels (Level, Division, Group, Class, Series), CPI value, date label, and trend indicator. The table includes several features:

The **filter row** at the top allows users to filter values in each column. The **Buttons extension** enables exporting the table to formats such as CSV, Excel, PDF, and Print. The **ColReorder extension** allows users to rearrange column positions. The **Responsive extension** ensures that important columns remain visible while lower priority columns collapse into a details view on smaller screens.

Additional table options control the display behaviour. Pagination is enabled with default page length of 25 rows. The length menu allows users to change the number of rows shown. Automatic column width adjustment helps fit the content neatly.

Column definitions further refine the display. Key columns such as Level, Division, Series, CPI value, date label, and Trend are always visible. Group and Class are given lower priority so they can collapse on smaller screens. All columns are centre aligned, and the Trend column uses the custom JavaScript renderer to display coloured symbols.

Finally, a **footer note** explains the meaning of the trend symbols and shows the CPI base year and data source.

```{r}
#| code-fold: true
#| code-summary: "SHOW"


cpi_h_table <- div(
  compact_dt_css,
  style = "
    width: 100%;
    max-width: 1200px;
    margin: 10px auto;
    overflow-x: hidden;
    font-family: 'Segoe UI', Arial, sans-serif;
  ",

  # Title
  div(
    style = 'text-align: center; margin-bottom: 20px;',
    h2('Singapore Consumer Price Index',
       style = 'color: #2c3e50; margin-bottom: 5px;'),
    h4(paste('Latest Data:',
             format(max(cpi_h_display$`As of`), '%B %Y')),
       style = 'color: #7f8c8d; font-weight: normal; margin-top: 0;')
  ),

  # Summary bar
  div(
    style = 'display: flex; justify-content: space-between; margin-bottom: 15px; padding: 10px; background-color: #f8f9fa; border-radius: 5px;',
    div(
      style = 'text-align: center; flex: 1;',
      div(style = 'font-size: 12px; color: #7f8c8d;', 'Total Rows'),
      div(style = 'font-size: 24px; font-weight: bold; color: #2c3e50;',
          nrow(cpi_h_display))
    ),
    div(
      style = 'text-align: center; flex: 1; border-left: 1px solid #dee2e6; border-right: 1px solid #dee2e6;',
      div(style = 'font-size: 12px; color: #7f8c8d;', 'Divisions'),
      div(style = 'font-size: 24px; font-weight: bold; color: #2c3e50;',
          cpi_h_display %>%
      filter(Level == "Major Category") %>%
      summarise(n = n_distinct(Division)) %>%
      pull(n))
    ),
    div(
      style = 'text-align: center; flex: 1;',
      div(style = 'font-size: 12px; color: #7f8c8d;', 'Last Updated'),
      div(style = 'font-size: 16px; font-weight: bold; color: #2c3e50;',
          format(max(cpi_h_display$`As of`), '%b %Y'))
    )
  ),

  # Main table
  datatable(
    # no more original_* columns to drop
    cpi_h_display %>%
      select(Level, Division, Group, Class, Series,
             `CPI Value`, `As of label`, Trend),
    class = 'cell-border stripe hover',
    filter = 'top',
    extensions = c('Buttons', 'ColReorder', 'Responsive'),
    width = '100%',
    options = list(
      pageLength = 25,
      lengthMenu = list(c(10, 25, 50, 100, -1),
                        c('10', '25', '50', '100', 'All')),
      dom = 'B<"toolbar">lfrtip',
      buttons = list(
        list(
          extend = 'collection',
          text   = 'Export',
          buttons = list(
            list(extend = 'copy', text = 'Copy'),
            list(extend = 'csv',  text = 'CSV'),
            list(extend = 'excel', text = 'Excel'),
            list(extend = 'pdf',  text = 'PDF'),
            list(extend = 'print', text = 'Print')
          )
        ),
        list(extend = 'colvis', text = 'Columns')
      ),
      scrollX = FALSE,
      scrollY = FALSE,
      autoWidth = TRUE,
      responsive = TRUE,
      columnDefs = list(
        # Always show: Level, Division, Series, CPI Value, As of label, Trend
        # column order: 0=Level,1=Division,2=Group,3=Class,4=Series,5=CPI,6=As of label,7=Trend
        list(className = 'all', targets = c(0, 1, 4, 5, 6, 7)),
        # Group & Class lower priority
        list(className = 'min-tablet', targets = c(2, 3)),
        # Center text
        list(className = 'dt-center', targets = '_all'),
        # Trend renderer
        list(
          targets  = 7,
          render = trend_js,
          orderable = FALSE,
          searchable = FALSE
        )
      )
    ),
    selection = 'single',
    rownames  = FALSE
  ),

  # Footer
  div(
    style = 'margin-top: 15px; font-size: 12px; color: #7f8c8d; text-align: center;',
    '▲ = Increase from previous month | ▼ = Decrease from previous month | ◆ = No change', br(),
    'Base Year: 2024 | Source: Singapore Department of Statistics & CEIC Database'
  )
)
```
:::

Here we have an **interactive CPI dashboard** that allows users to filter, explore hierarchy levels, export data, and interpret CPI movements.

```{r}
cpi_h_table
```


```{r}
glimpse(cpi_h)
```


## EDA

### Trend

#### Top Level CPI 

##### Single View 



```{r}
#| code-fold: true
#| code-summary: "SHOW"
create_cpi_plot <- function(data, 
                            plot_level,
                            category = NULL,
                            start_date = "2020-01-01",
                            end_date = "2026-01-01",
                            plot_title = NULL,
                            ref_line_color = "black",
                            ref_line_width = 0.9,
                            plot_line_width = 0.6) {
  
  # Validate and format dates
  start_date <- as.Date(start_date)
  end_date <- as.Date(end_date)
  
  if (start_date >= end_date) {
    stop("start_date must be before end_date")
  }
  
  # Setup based on level
  if (plot_level == 0) {
    plot_data <- data %>% filter(level == 0)
    ref_data <- NULL
    ref_label <- NULL
    default_title <- "Signapore CPI Trend"
    color_label <- "Index"
    
  } else if (plot_level == 1) {
    plot_data <- data %>% filter(level == 1)
    ref_data <- data %>% filter(level == 0, series == "Consumer Price Index (CPI)")
    ref_label <- "Overall CPI"
    default_title <- "CPI Major Groups"
    color_label <- "Category"
    
  } else if (plot_level == 2) {
    if (is.null(category)) stop("For level 2, specify a category (division name)")
    plot_data <- data %>% filter(level == 2, division == category)
    ref_data <- data %>% filter(level == 1, series == category)
    ref_label <- category
    default_title <- paste(category, ": Subgroups")
    color_label <- "Subcategory"
    
  } else if (plot_level == 3) {
    if (is.null(category)) stop("For level 3, specify a group")
    plot_data <- data %>% filter(level == 3, group == category)
    ref_data <- data %>% filter(level == 2, series == category)
    ref_label <- category
    default_title <- paste(category, ": Detailed Classes")
    color_label <- "Detailed Class"
    
  } else {
    stop("plot_level must be 0, 1, 2, or 3")
  }
  
  # Check if data exists after filtering
  if (nrow(plot_data) == 0) {
    stop(paste0("No data found for plot_level=", plot_level, 
                ifelse(!is.null(category), paste0(" and category='", category, "'"), "")))
  }
  
  # Use custom or default title
  if (is.null(plot_title)) {
    plot_title <- default_title
  }
  
  has_reference <- !is.null(ref_data) && nrow(ref_data) > 0
  
  # Calculate date breaks safely
  date_breaks_seq <- seq(start_date, end_date, by = "6 months")
  if (length(date_breaks_seq) < 2) {
    date_breaks_seq <- seq(start_date, end_date, by = "2 months")
  }
  
  # Build ggplot
  p <- ggplot(plot_data, aes(x = date, y = cpi, color = series)) +
    geom_line(linewidth = plot_line_width, alpha = 0.7)
  
  if (has_reference) {
    p <- p + geom_line(data = ref_data, aes(x = date, y = cpi),
                       color = ref_line_color, linewidth = ref_line_width,
                       linetype = "solid", inherit.aes = FALSE)
  }
  
  p <- p + labs(
    title = plot_title,
    subtitle = if (has_reference) paste("Black line =", ref_label) else "",
    x = "Date", 
    y = "CPI Index", 
    color = color_label
  ) +
  scale_x_date(
    limits = c(start_date, end_date),
    breaks = date_breaks_seq,  # Use pre-calculated breaks
    date_labels = "%b%y",
    expand = expansion(mult = c(0.01, 0.01)) 
  ) +
  theme_minimal() +
  theme(
    legend.position = "bottom",
    axis.text.x = element_text(hjust = 1),
    plot.title = element_text(face = "bold", size = 14, hjust = 0.5),
    plot.subtitle = element_text(size = 10, hjust = 0.5, color = "gray40")
  )
  
  # Convert to plotly
  ggplotly(p) %>%
    layout(
      title = list(
         text = paste0("<b>", plot_title, ", 2020-2025</b>"), 
        font = list(size = 16, family = "Arial", color = "#2c3e50"),
        x = 0.5, 
        xanchor = "center"
      ),
      legend = list(
        orientation = "h", 
        y = -0.3, 
        x = 0.5, 
        xanchor = "center",
        font = list(family = "Arial", size = 9, color = "#333333"),
        itemsizing = "constant", 
        itemwidth = 35,
        bgcolor = "rgba(255,255,255,0.9)",
        bordercolor = "#cccccc", 
        borderwidth = 1
      ),
      annotations = list(list(
        text = paste0(
          "<b>Source:</b> CEIC Database; Index, 2024 = 100",
          if (has_reference) paste0("; Black line = ", ref_label) else ""
        ),
        x = 1, y = -0.25, 
        xref = "paper", yref = "paper", 
        showarrow = FALSE,
        font = list(family = "Arial", size = 9, color = "#666666"),
        xanchor = "right", 
        align = "right"
      )),
      xaxis = list(
        tickfont = list(size = 9), 
        gridcolor = "#f0f0f0", 
        tickangle = 0
      ),
      yaxis = list(
        tickfont = list(size = 9), 
        gridcolor = "#f0f0f0", 
        zerolinecolor = "#cccccc"
      ),
      hoverlabel = list(
        bgcolor = "white", 
        bordercolor = "#cccccc", 
        font = list(size = 10, family = "Arial")
      ),
      margin = list(b = 150, l = 60, r = 40, t = 60),
      plot_bgcolor = "white", 
      paper_bgcolor = "white"
    )
}

```



```{r}
create_cpi_plot(cpi_h, plot_level = 0)
```


##### Facet View
```{r}
plot_cpi_time_series <- function(data,
                                 plot_level,
                                 category = NULL,
                                 group_name = NULL,
                                 add_year = TRUE,
                                 use_year_colour = FALSE,
                                 facet_by = TRUE,
                                 facet_ncol = 1,
                                 facet_nrow = 1,
                                 facet_scales = "free_y",
                                 facet_dir = "h",
                                 facet_collapse = FALSE,
                                 facet_collapse_sep = " ",
                                 facet_strip_remove = FALSE,
                                 line_color = "#2c3e50",
                                 line_size = 0.5,
                                 line_type = 1,
                                 line_alpha = 1,
                                 y_intercept = NULL,
                                 y_intercept_color = "#2c3e50",
                                 x_intercept = NULL,
                                 x_intercept_color = "#2c3e50",
                                 smooth = FALSE,
                                 smooth_period = "auto",
                                 smooth_message = FALSE,
                                 smooth_span = NULL,
                                 smooth_degree = 2,
                                 smooth_color = "#3366FF",
                                 smooth_size = 1,
                                 smooth_alpha = 1,
                                 legend_show = FALSE,
                                 title = NULL,
                                 x_lab = "",
                                 y_lab = "CPI Index",
                                 color_lab = "Legend",
                                 interactive = TRUE,
                                 plotly_slider = FALSE,
                                 trelliscope = FALSE,
                                 trelliscope_params = list(),
                                 return_plot = FALSE) {
  
  if (add_year) {
    data <- data %>%
      dplyr::mutate(year = lubridate::year(date))
  }
  
  if (plot_level == 0) {
    plot_data <- data %>% dplyr::filter(level == 0)
    default_title <- "CPI Top Level Time Series"
    facet_col <- NULL
    
  } else if (plot_level == 1) {
    plot_data <- data %>% dplyr::filter(level == 1)
    default_title <- "CPI Major Groups Time Series"
    facet_col <- "division"
    
  } else if (plot_level == 2) {
    if (is.null(category)) stop("For level 2, specify a category")
    
    plot_data <- data %>%
      dplyr::filter(level == 2, division == category)
    
    default_title <- paste(category, "CPI Sub Categories Time Series")
    facet_col <- "series"
    
  } else if (plot_level == 3) {
    if (is.null(group_name)) stop("For level 3, specify a group_name")
    
    plot_data <- data %>%
      dplyr::filter(level == 3, group == group_name)
    
    default_title <- paste(group_name, "Detailed CPI Time Series")
    facet_col <- "series"
    
  } else {
    stop("plot_level must be 0, 1, 2, or 3")
  }
  
  if (nrow(plot_data) == 0) {
    stop("No data found for the specified criteria")
  }
  
  if (is.null(title)) {
    title <- default_title
  }
  
  # Build the call safely based on user options
  if (use_year_colour && facet_by && !is.null(facet_col) && "year" %in% names(plot_data)) {
    
    p <- timetk::plot_time_series(
      .data = plot_data,
      .date_var = date,
      .value = cpi,
      .color_var = year,
      .facet_vars = tidyselect::all_of(facet_col),
      .facet_ncol = facet_ncol,
      .facet_nrow = facet_nrow,
      .facet_scales = facet_scales,
      .facet_dir = facet_dir,
      .facet_collapse = facet_collapse,
      .facet_collapse_sep = facet_collapse_sep,
      .facet_strip_remove = facet_strip_remove,
      .line_color = line_color,
      .line_size = line_size,
      .line_type = line_type,
      .line_alpha = line_alpha,
      .y_intercept = y_intercept,
      .y_intercept_color = y_intercept_color,
      .x_intercept = x_intercept,
      .x_intercept_color = x_intercept_color,
      .smooth = smooth,
      .smooth_period = smooth_period,
      .smooth_message = smooth_message,
      .smooth_span = smooth_span,
      .smooth_degree = smooth_degree,
      .smooth_color = smooth_color,
      .smooth_size = smooth_size,
      .smooth_alpha = smooth_alpha,
      .legend_show = legend_show,
      .title = title,
      .x_lab = x_lab,
      .y_lab = y_lab,
      .color_lab = color_lab,
      .interactive = interactive,
      .plotly_slider = plotly_slider,
      .trelliscope = trelliscope,
      .trelliscope_params = trelliscope_params
    )
    
  } else if (use_year_colour && !facet_by && "year" %in% names(plot_data)) {
    
    p <- timetk::plot_time_series(
      .data = plot_data,
      .date_var = date,
      .value = cpi,
      .color_var = year,
      .facet_ncol = facet_ncol,
      .facet_nrow = facet_nrow,
      .facet_scales = facet_scales,
      .facet_dir = facet_dir,
      .facet_collapse = facet_collapse,
      .facet_collapse_sep = facet_collapse_sep,
      .facet_strip_remove = facet_strip_remove,
      .line_color = line_color,
      .line_size = line_size,
      .line_type = line_type,
      .line_alpha = line_alpha,
      .y_intercept = y_intercept,
      .y_intercept_color = y_intercept_color,
      .x_intercept = x_intercept,
      .x_intercept_color = x_intercept_color,
      .smooth = smooth,
      .smooth_period = smooth_period,
      .smooth_message = smooth_message,
      .smooth_span = smooth_span,
      .smooth_degree = smooth_degree,
      .smooth_color = smooth_color,
      .smooth_size = smooth_size,
      .smooth_alpha = smooth_alpha,
      .legend_show = legend_show,
      .title = title,
      .x_lab = x_lab,
      .y_lab = y_lab,
      .color_lab = color_lab,
      .interactive = interactive,
      .plotly_slider = plotly_slider,
      .trelliscope = trelliscope,
      .trelliscope_params = trelliscope_params
    )
    
  } else if (!use_year_colour && facet_by && !is.null(facet_col)) {
    
    p <- timetk::plot_time_series(
      .data = plot_data,
      .date_var = date,
      .value = cpi,
      .facet_vars = tidyselect::all_of(facet_col),
      .facet_ncol = facet_ncol,
      .facet_nrow = facet_nrow,
      .facet_scales = facet_scales,
      .facet_dir = facet_dir,
      .facet_collapse = facet_collapse,
      .facet_collapse_sep = facet_collapse_sep,
      .facet_strip_remove = facet_strip_remove,
      .line_color = line_color,
      .line_size = line_size,
      .line_type = line_type,
      .line_alpha = line_alpha,
      .y_intercept = y_intercept,
      .y_intercept_color = y_intercept_color,
      .x_intercept = x_intercept,
      .x_intercept_color = x_intercept_color,
      .smooth = smooth,
      .smooth_period = smooth_period,
      .smooth_message = smooth_message,
      .smooth_span = smooth_span,
      .smooth_degree = smooth_degree,
      .smooth_color = smooth_color,
      .smooth_size = smooth_size,
      .smooth_alpha = smooth_alpha,
      .legend_show = legend_show,
      .title = title,
      .x_lab = x_lab,
      .y_lab = y_lab,
      .color_lab = color_lab,
      .interactive = interactive,
      .plotly_slider = plotly_slider,
      .trelliscope = trelliscope,
      .trelliscope_params = trelliscope_params
    )
    
  } else {
    
    p <- timetk::plot_time_series(
      .data = plot_data,
      .date_var = date,
      .value = cpi,
      .facet_ncol = facet_ncol,
      .facet_nrow = facet_nrow,
      .facet_scales = facet_scales,
      .facet_dir = facet_dir,
      .facet_collapse = facet_collapse,
      .facet_collapse_sep = facet_collapse_sep,
      .facet_strip_remove = facet_strip_remove,
      .line_color = line_color,
      .line_size = line_size,
      .line_type = line_type,
      .line_alpha = line_alpha,
      .y_intercept = y_intercept,
      .y_intercept_color = y_intercept_color,
      .x_intercept = x_intercept,
      .x_intercept_color = x_intercept_color,
      .smooth = smooth,
      .smooth_period = smooth_period,
      .smooth_message = smooth_message,
      .smooth_span = smooth_span,
      .smooth_degree = smooth_degree,
      .smooth_color = smooth_color,
      .smooth_size = smooth_size,
      .smooth_alpha = smooth_alpha,
      .legend_show = legend_show,
      .title = title,
      .x_lab = x_lab,
      .y_lab = y_lab,
      .color_lab = color_lab,
      .interactive = interactive,
      .plotly_slider = plotly_slider,
      .trelliscope = trelliscope,
      .trelliscope_params = trelliscope_params
    )
  }
  
  if (return_plot) {
    return(p)
  } else {
    print(p)
  }
}
```


```{r}
plot_cpi_time_series(
  cpi_h,
  plot_level = 1,
  use_year_colour = TRUE,
  facet_by = TRUE,
  facet_ncol = 5,  interactive = FALSE
)
```





```{r}
#| code-fold: true
#| code-summary: "SHOW"

p <- cpi_h %>%
  filter(level == 0) %>%
  group_by(series) %>%
  plot_time_series(
    .date_var = date,
    .value = cpi,
    .color_var = series,
    .facet_ncol = 3,
    .facet_scales = "free_y",
    .interactive = TRUE,
    .title = "Singapore CPI Trend, 2020 – 2025",
    .y_lab = "CPI Index",
    .x_lab = "Date",
    .plotly_slider = FALSE,
    .smooth = TRUE
  )

p %>% layout(
  legend = list(
    orientation = "h", # Horizontal
    y = -0.25,         # Below plot
    x = 0.5,              # Center
    xanchor = "center"
  ),
  annotations = list(
    list(
      text = "<b>Source:</b> CEIC Database; Index, 2024 = 100",
      x = 1,           # Right align caption
      y = -0.37,      # Below legend
      xref = "paper",
      yref = "paper",
      showarrow = FALSE,
      font = list(size = 10, color = "gray50"),
      xanchor = "right"
    )
  ),
  margin = list(b = 120)  # Extra bottom space
)

```

#### Major Group

##### Single View

```{r}
create_cpi_plot(cpi_h, plot_level = 1)
```



##### Facet View


```{r}
#| code-fold: true
#| code-summary: "SHOW"
make_cpi_plot_improved <- function(data, level_val, 
                          category_filter = NULL, 
                          group_filter = NULL, interactive = FALSE,
                          smooth = FALSE, facet_ncol = 3) {
  
  # Apply category filter (division level) if specified
  if (!is.null(category_filter)) {
    data <- data %>%
      filter(division == category_filter)
  }
  
  # Apply group filter if specified
  if (!is.null(group_filter)) {
    data <- data %>%
      filter(group == group_filter)
  }
  
  # Determine title
  if (!is.null(group_filter)) {
    base_title <- paste(group_filter, "CPI")
  } else if (!is.null(category_filter)) {
    base_title <- paste(category_filter, "CPI")
  } else {
    base_title <- "CPI"
  }
  
  if (level_val == 1) {
    # Level 1: Major groups - facet by division
    data %>%
      filter(level == 1) %>%
      group_by(division) %>%
      plot_time_series(
        .date_var = date,
        .value = cpi,
        .facet_ncol = facet_ncol,
        .facet_scales = "free_y",
        .smooth = smooth,
        .interactive = interactive,
        .title = paste(base_title, "- Major Groups (Level 1)"),
        .y_lab = "CPI Index"
        
        
      )
      
  } else if (level_val == 2) {
    # Level 2: Sub-groups - facet by group
    data %>%
      filter(level == 2) %>%
      group_by(group) %>%
      plot_time_series(
        .date_var = date,
        .value = cpi,
        .facet_ncol = facet_ncol,
        .facet_scales = "free_y",
        .smooth = smooth,
        .interactive = interactive,
        .title = paste(base_title, "- Subgroups (Level 2)"),
        .y_lab = "CPI Index"
      )
      
  } else if (level_val == 3) {
    # Level 3: Detailed classes - facet by class
    data %>%
      filter(level == 3) %>%
      group_by(class) %>%
      plot_time_series(
        .date_var = date,
        .value = cpi,
        .facet_ncol = facet_ncol,
        .facet_scales = "free_y",
        .smooth = smooth,
        .interactive = interactive,
        .title = paste(base_title, "- Detailed Classes (Level 3)"),
        .y_lab = "CPI Index"
      )
      
  } else {
    stop("level_val must be 1, 2, or 3")
  }
}

```


```{r}
make_cpi_plot_improved(cpi_h, level = 1)
```

```{r}
unique(cpi_h$series[cpi_h$level == 1])
```

:::panel-tabset

#### Food

##### Subgroup 

```{r}
unique(cpi_h$series[cpi_h$division == "Food" & cpi_h$level == 2])
```


###### Single View
```{r}
# Level 2: Food subgroups with Food as reference
create_cpi_plot(cpi_h, plot_level = 2, category = "Food")
```


###### Facet View
```{r}
make_cpi_plot_improved(cpi_h, level_val = 2, category_filter = "Food")
```


##### FFBSS

"Food Excl Food & Beverage Serving Services (FFBSS)"
[2] "Food & Beverage Serving Services (FBSS)"    

###### Single View

```{r}

# Level 3: FFBSS detailed classes with FFBSS as reference
create_cpi_plot(cpi_h, plot_level = 3, category = "Food Excl Food & Beverage Serving Services (FFBSS)")
```




###### Facet View

```{r}
make_cpi_plot_improved(cpi_h, level_val = 3, 
      category_filter = "Food",group_filter = "Food & Beverage Serving Services (FBSS)",facet_ncol = 4)
```



##### FBSS

###### Single View

```{r}
create_cpi_plot(cpi_h, plot_level = 3, category = "Food & Beverage Serving Services (FBSS)")

```


###### Facet View

```{r}
make_cpi_plot_improved(cpi_h, level_val = 3, category_filter = "Food",group_filter = "Food & Beverage Serving Services (FBSS)",facet_ncol = 2)
```



#### Clothing and Footwear (C&F)

##### Single View

```{r}
create_cpi_plot(cpi_h, plot_level = 2, category = "Clothing and Footwear (C&F)")
```


##### Facet View

```{r}
make_cpi_plot_improved(cpi_h, level_val = 2, 
category_filter = "Clothing and Footwear (C&F)")
```


#### Housing & Utilities



##### Single View

```{r}
create_cpi_plot(cpi_h, plot_level = 2, category = "Housing & Utilities")
```

##### Facet View

```{r}
make_cpi_plot_improved(cpi_h, level_val = 2,
category_filter = "Housing & Utilities")
```

#### Household Durables & Services (HDS)

##### Single View

```{r}
create_cpi_plot(cpi_h, plot_level = 2, category = "Household Durables & Services (HDS)")
```

##### Facet View

```{r}
make_cpi_plot_improved(cpi_h, level_val = 2, category_filter = "Household Durables & Services (HDS)")
```


#### Health

##### Single View

```{r}
create_cpi_plot(cpi_h, plot_level = 2, category = "Health")

```


##### Facet View

```{r}
make_cpi_plot_improved(cpi_h, level_val = 2, category_filter = "Health")
```

#### Transport

##### Single View

```{r}
create_cpi_plot(cpi_h, plot_level = 2, category = "Transport")

```

##### Facet View

```{r}
make_cpi_plot_improved(cpi_h, level_val = 2, category_filter = "Transport", facet_ncol = 2)
```

#### Information & Communication (InfoComm)


##### Single View

```{r}
create_cpi_plot(cpi_h, plot_level = 2, category = "Information & Communication (InfoComm)")
```

##### Facet View

```{r}
make_cpi_plot_improved(cpi_h, level_val = 2, category_filter = "Information & Communication (InfoComm)")
```



#### Recreation, Sport & Culture (RSC)   

##### Single View

```{r}
create_cpi_plot(cpi_h, plot_level = 2, category = "Recreation, Sport & Culture (RSC)")
```


##### Facet View

```{r}
make_cpi_plot_improved(cpi_h, level_val = 2, category_filter = "Recreation, Sport & Culture (RSC)")
```


#### Education

##### Single View

```{r}
create_cpi_plot(cpi_h, plot_level = 2, category = "Education")
```


##### Facet View

```{r}
make_cpi_plot_improved(cpi_h, level_val = 2, category_filter = "Education")
```

#### Miscellaneous Goods & Services (MG&S)

##### Single View

```{r}
create_cpi_plot(cpi_h, plot_level = 2, category = "Miscellaneous Goods & Services (MG&S)")
```


##### Facet View

```{r}
make_cpi_plot_improved(cpi_h, level_val = 2, category_filter = "Miscellaneous Goods & Services (MG&S)")
```

:::




```{r}
#| eval: false
library(shiny)
library(plotly)  

ui <- fluidPage(
  titlePanel("Singapore CPI explorer"),
  sidebarLayout(
    sidebarPanel(
      selectInput(
        "level", "Detail level",
        choices = c("Divisions" = 1, "Groups" = 2, "Classes" = 3),
        selected = 1
      )
    ),
    mainPanel(
      plotlyOutput("cpi_plot", height = "700px")
    )
  )
)

server <- function(input, output, session) {
  output$cpi_plot <- renderPlotly({
    make_cpi_plot(cpi_h, as.integer(input$level))
  })
}
shinyApp(ui, server)
```


### Seasonality


#### CPI All

##### Cycle Plot

```{r}
#| code-fold: true
#| code-summary: "SHOW"

cpi_cycle <- cpi_h %>%
  # Filter for overall CPI or main category of interest
  filter(series == "Consumer Price Index (CPI)") %>%
  mutate(
    year = year(date),
    month = month(date, label = TRUE, abbr = TRUE),
    month_num = month(date),
    # Create a year label for coloring
    year_label = as.factor(year)
  ) %>%
  # Ensure we have complete years
  group_by(year) %>%
  filter(n() >= 12) %>%
  ungroup()

## 1. Exact replica of example image
exact_replica_plot <- ggplot(cpi_cycle, aes(x = month, y = cpi, group = year, color = year_label)) +
  geom_line(size = 0.8) +
  geom_point(size = 1.5) +
  # Use a blue gradient similar to the image
  scale_color_manual(
    values = c(
      "2020" = "#D4E6F1",
      "2021" = "#A9CCE3",
      "2022" = "#7FB3D5",
      "2023" = "#5499C7",
      "2024" = "#2E86C1",
      "2025" = "#1B4F72"
    ),
    name = "Year"
  ) +
  labs(
    title = "Consumer Price Index (CPI) by Month, 2020-2025",
    x = "",
    y = "CPI Value"
  ) +
  theme_minimal() +
  theme(
    # Remove grid lines except horizontal
    panel.grid.major.x = element_blank(),
    panel.grid.minor.x = element_blank(),
    panel.grid.major.y = element_line(color = "gray90", size = 0.3),
    panel.grid.minor.y = element_blank(),
    # Clean background
    panel.background = element_rect(fill = "white", color = NA),
    plot.background = element_rect(fill = "white", color = NA),
    # Axis lines
    axis.line = element_line(color = "black", size = 0.3),
    axis.ticks = element_line(color = "black"),
    axis.ticks.length = unit(0.2, "cm"),
    # Text formatting
    axis.text.x = element_text(angle = 0, hjust = 0.5, size = 10, color = "black"),
    axis.text.y = element_text(size = 9, color = "black"),
    axis.title.y = element_text(size = 11, face = "plain", margin = margin(r = 10)),
    # Title formatting
    plot.title = element_text(size = 14, face = "bold", hjust = 0.5, margin = margin(b = 10)),
    # Legend
    legend.position = "right",
    legend.title = element_text(size = 10, face = "bold"),
    legend.text = element_text(size = 9),
    # Remove all other grid lines
    panel.border = element_blank()
  ) +
  # Ensure months are in correct order
  scale_x_discrete(limits = month.abb) +
  # Format y-axis with comma for thousands if needed
  scale_y_continuous(labels = scales::number_format(accuracy = 0.1))

# Display the plot
exact_replica_plot
```



##### Sesonal Plot


```{r}
#| code-fold: true
#| code-summary: "SHOW"

cpi_h %>%
  filter(series == "Consumer Price Index (CPI)") %>%
  plot_seasonal_diagnostics(
    .date_var = date,
    .value = cpi,
    .feature_set = c("month.lbl", "quarter", "year"),
    .geom = "boxplot",
    .geom_color = "scale_color",
    .title = "Seasonal Diagnostics: Consumer Price Index (CPI)",
    .interactive = FALSE
  )
```


#### Major Groups

##### Calender Heatmap

```{r}
#| code-fold: true
#| code-summary: "SHOW"

cpi_level1 <- cpi_h %>%
  filter(level == 1)

start_date <- min(cpi_level1$date)
end_date <- max(cpi_level1$date)
date_breaks_seq <- seq(start_date, end_date, by = "6 months")

if (length(date_breaks_seq) < 2) {
  date_breaks_seq <- seq(start_date, end_date, by = "2 months")
}

p <- ggplot(cpi_level1, aes(x = date, y = series, fill = cpi,
                            text = paste0("Major Group: ", series,
                                          "<br>Date: ", format(date, "%b %Y"),
                                          "<br>CPI: ", round(cpi, 2)))) +
  geom_tile(color = "white", linewidth = 0.2) +
  scale_fill_viridis_c(
    option = "viridis",  
    direction = -1,    
    labels = number_format()
  ) +
  scale_x_date(
    limits = c(start_date, end_date),
    breaks = date_breaks_seq,
    date_labels = "%b%y",
    expand = expansion(mult = c(0.01, 0.01))
  ) +
  labs(
    title = "CPI Level by Major Group",
    x = "Date",
    y = "Group",
    fill = "CPI Index"
  ) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1, size = 7),
    axis.text.y = element_text(size = 7),
    axis.title = element_text(size = 8),
    legend.position = "bottom",
    legend.title = element_text(size = 8),
    legend.text = element_text(size = 7),
    panel.grid = element_blank()
  )

ggplotly(p, tooltip = "text") %>%
  layout(
    title = list(
      text = paste0("<b>CPI Level by Category</b><br>",
                    "<sup>CPI Major Groups, ", format(start_date, "%Y"), "-", format(end_date, "%Y"), "</sup>"),
      font = list(size = 14, family = "Arial", color = "#2c3e50"),
      x = 0.5, xanchor = "center"
    ),
    xaxis = list(
      tickangle = 0, 
      tickfont = list(size = 7), 
      titlefont = list(size = 8),
      gridcolor = "#f0f0f0"
    ),
    yaxis = list(
      tickfont = list(size = 7), 
      titlefont = list(size = 8),
      gridcolor = "#f0f0f0"
    ),
    legend = list(
      orientation = "h", 
      y = -0.3, 
      x = 0.5, 
      xanchor = "center",
      title = list(text = "CPI Index", font = list(size = 8)),
      font = list(size = 7)
    ),
    annotations = list(list(
      text = "<b>Source:</b> CEIC Database; Index, 2024 = 100",
      x = 1, y = -0.22, xref = "paper", yref = "paper", showarrow = FALSE,
      font = list(family = "Arial", size = 8, color = "#666666"),
      xanchor = "right", align = "right"
    )),
    margin = list(b = 120, l = 120, r = 40, t = 60),
    plot_bgcolor = "white", 
    paper_bgcolor = "white"
  )

```

##### Function 


```{r}
#| code-fold: true
#| code-summary: "SHOW"
plot_cpi_seasonal <- function(data, 
                              plot_level,
                              category = NULL,
                              group = NULL,
                              interactive = FALSE,
                              geom = "boxplot",
                              geom_color = "#2c3e50",
                              feature_set = c("month.lbl", "quarter", "year"),
                              return_plots = FALSE,
                              pause_between = FALSE) {
  
  # Determine series list based on level
  if (plot_level == 0) {
    series_list <- data %>% 
      filter(level == 0) %>% 
      distinct(series) %>% 
      pull(series)
      
  } else if (plot_level == 1) {
    series_list <- data %>% 
      filter(level == 1) %>% 
      distinct(series) %>% 
      pull(series)
      
  } else if (plot_level == 2) {
    if (is.null(category)) stop("For level 2, specify a category (division name)")
    series_list <- data %>% 
      filter(level == 2, division == category) %>% 
      distinct(series) %>% 
      pull(series)
      
  } else if (plot_level == 3) {
    if (is.null(group)) stop("For level 3, specify a group")
    series_list <- data %>% 
      filter(level == 3, group == group) %>% 
      distinct(series) %>% 
      pull(series)
      
  } else {
    stop("plot_level must be 0, 1, 2, or 3")
  }
  
  if (length(series_list) == 0) {
    stop("No series found for the specified criteria")
  }
  
  cat("Plotting", length(series_list), "series\n")
  
  plots <- list()
  
  for (s in series_list) {
    plot_data <- data %>% filter(series == s)
    
    p <- plot_seasonal_diagnostics(
      .data = plot_data,
      .date_var = date,
      .value = cpi,
      .feature_set = feature_set,
      .geom = geom,
      .geom_color = geom_color,
      .title = paste("Seasonal Diagnostics:", s),
      .x_lab = "",
      .y_lab = "",
      .interactive = interactive
    )
    
    if (return_plots) {
      plots[[s]] <- p
    } else {
      print(p)
    }
    
    if (pause_between && s != tail(series_list, 1)) {
      readline(prompt = "Press Enter for next plot...")
    }
  }
  
  if (return_plots) invisible(plots)
}

```

```{r}
plot_cpi_seasonal(cpi_h, 0, interactive = FALSE)
```


```{r}
plot_cpi_seasonal(cpi_h, 1,interactive = FALSE)
```




```{r}
plot_cpi_seasonal(cpi_h, 2, category = "Food", interactive = FALSE)
```


```{r}
plot_cpi_seasonal(cpi_h, 3, group = "FFBSS", interactive = FALSE)
```



### Autocorrelation




```{r}
#| eval: false
output$acf_plot <- renderPlotly({
  cpi_h %>%
    filter(series == input$acf_series) %>%  # e.g. 'Consumer Price Index (CPI)'
    plot_acf_diagnostics(
      date, cpi,
      .lags = 36,
      .interactive = FALSE
    )
})



```



```{r}
#| code-fold: true
#| code-summary: "SHOW"

plot_cpi_acf <- function(data,
                         plot_level,
                         category = NULL,
                         group_name = NULL,
                         lags = 36,
                         interactive = FALSE,
                         facet_ncol = 3,
                         facet_scales = "fixed",
                         line_color = "#2c3e50",
                         line_size = 0.5,
                         line_alpha = 1,
                         point_color = "#2c3e50",
                         point_size = 1,
                         point_alpha = 1,
                         x_intercept = NULL,
                         x_intercept_color = "#E31A1C",
                         hline_color = "#2c3e50",
                         white_noise_line_type = 2,
                         white_noise_line_color = "#A6CEE3",
                         show_ccf_vars_only = FALSE,
                         show_white_noise_bars = TRUE,
                         plotly_slider = FALSE,
                         return_plot = FALSE,
                         ccf_target = NULL,
                         ccf_predictors = NULL) {
  
  
  # Step 1: filter hierarchy
  if (plot_level == 0) {
    plot_data <- data %>% filter(level == 0)
    plot_title <- "CPI Top Level: Lag Diagnostics"
    
  } else if (plot_level == 1) {
    plot_data <- data %>% filter(level == 1)
    plot_title <- "CPI Major Groups: Lag Diagnostics"
    
  } else if (plot_level == 2) {
    if (is.null(category)) {
      stop("For level 2, specify a category")
    }
    plot_data <- data %>% filter(level == 2, division == category)
    plot_title <- paste(category, ": Sub-Categories Lag Diagnostics")
    
  } else if (plot_level == 3) {
    if (is.null(group_name)) {
      stop("For level 3, specify a group_name")
    }
    plot_data <- data %>% filter(level == 3, group == group_name)
    plot_title <- paste(group_name, ": Detailed Classes Lag Diagnostics")
    
  } else {
    stop("plot_level must be 0, 1, 2, or 3")
  }
  
  if (nrow(plot_data) == 0) {
    stop("No data found for the specified criteria")
  }
  
  use_ccf <- !is.null(ccf_target) && !is.null(ccf_predictors)
  
  # Step 2A: ACF/PACF mode
  if (!use_ccf) {
    
    if (plot_level == 0) {
      acf_data <- plot_data
    } else if (plot_level == 1) {
      acf_data <- plot_data %>% group_by(division)
    } else {
      acf_data <- plot_data %>% group_by(series)
    }
    
    p <- timetk::plot_acf_diagnostics(
      .data = acf_data,
      .date_var = date,
      .value = cpi,
      .lags = lags,
      .show_white_noise_bars = show_white_noise_bars,
      .facet_ncol = facet_ncol,
      .facet_scales = facet_scales,
      .line_color = line_color,
      .line_size = line_size,
      .line_alpha = line_alpha,
      .point_color = point_color,
      .point_size = point_size,
      .point_alpha = point_alpha,
      .x_intercept = x_intercept,
      .x_intercept_color = x_intercept_color,
      .hline_color = hline_color,
      .white_noise_line_type = white_noise_line_type,
      .white_noise_line_color = white_noise_line_color,
      .title = plot_title,
      .x_lab = "Lag",
      .y_lab = "Correlation",
      .interactive = interactive,
      .plotly_slider = plotly_slider
    )
    
  } else {
    
    # Step 2B: CCF mode
    series_needed <- c(ccf_target, ccf_predictors)
    
    missing_series <- setdiff(series_needed, unique(plot_data$series))
    if (length(missing_series) > 0) {
      stop(
        paste(
          "These series were not found in the filtered data:",
          paste(missing_series, collapse = ", ")
        )
      )
    }
    
    wide_data <- plot_data %>%
      filter(series %in% series_needed) %>%
      select(date, series, cpi) %>%
      distinct() %>%
      pivot_wider(names_from = series, values_from = cpi) %>%
      arrange(date) %>%
      drop_na(all_of(series_needed))
    
    if (nrow(wide_data) == 0) {
      stop("No complete cases available after reshaping to wide format.")
    }
    
    target_sym <- rlang::sym(ccf_target)
    
    p <- timetk::plot_acf_diagnostics(
      .data = wide_data,
      .date_var = date,
      .value = !!target_sym,
      .ccf_vars = all_of(ccf_predictors),
      .lags = lags,
      .show_ccf_vars_only = show_ccf_vars_only,
      .show_white_noise_bars = show_white_noise_bars,
      .facet_ncol = facet_ncol,
      .facet_scales = facet_scales,
      .line_color = line_color,
      .line_size = line_size,
      .line_alpha = line_alpha,
      .point_color = point_color,
      .point_size = point_size,
      .point_alpha = point_alpha,
      .x_intercept = x_intercept,
      .x_intercept_color = x_intercept_color,
      .hline_color = hline_color,
      .white_noise_line_type = white_noise_line_type,
      .white_noise_line_color = white_noise_line_color,
      .title = paste0(
        plot_title,
        " | Target: ", ccf_target,
        " | Predictors: ", paste(ccf_predictors, collapse = ", ")
      ),
      .x_lab = "Lag",
      .y_lab = "Correlation",
      .interactive = interactive,
      .plotly_slider = plotly_slider
    )
  }
  
  if (return_plot) {
    return(p)
  } else {
    print(p)
  }
}
```


```{r}
plot_cpi_acf(cpi_h, plot_level = 1)
```

```{r}
plot_cpi_acf(cpi_h, plot_level = 2, category = "Food")
```

```{r}
plot_cpi_acf(
  cpi_h,
  plot_level = 3,
  group_name = "Food Excl Food & Beverage Serving Services (FFBSS)",
  facet_ncol = 3
)
```



```{r}
plot_cpi_acf(
  cpi_h,
  plot_level = 3,
  group_name = "Food Excl Food & Beverage Serving Services (FFBSS)",
  ccf_target = "FFBSS: Ready-Made & Oth. Food Nec",
  ccf_predictors = c("FFBSS: Vegetables", "FFBSS: Meat"),
  lags = 24,
  facet_ncol = 1
)
```


