---
title: "04-4  Funnel Plots for Fair Comparisons"
author: "Yiqiong PAN"
date: "02 Feb, 2026"
date-modified: last-modified

execute:
  echo: true    # display the code
  eval: true    # run the code
  message: false
  warning: false
  freeze: true  # do not re-render if nothing edited

format: 
  html:
    toc: true
    code-link: true

number-sections: true
number-depth: 6

editor: visual
---

## Learning Outcome

A funnel plot is a specialised data visualisation used to support fair and unbiased comparisons across outlets, stores, or business entities. In this hands on exercise, we learn how to create funnel plots using the **funnelPlotR** package, build static funnel plots with **ggplot2**, and develop interactive funnel plots by combining **ggplot2** with **plotly**.

**Assumption:**

All hypothesis tests in this exercise are conducted at the **5% significance level**, where the null hypothesis states that no significant difference or no relationship exists between the variables of interest.

## Getting Started

### Installing and Loading the Required Libraries

There are four packages that will be used in this exercise.

| Package | Description |
|--------------------------|----------------------------------------------|
| [tidyverse](https://www.tidyverse.org/) | A collection of R packages for data importing, cleaning, transformation, and visualisation (e.g., [readr](https://readr.tidyverse.org/), dplyr, tidyr, [ggplot2](https://ggplot2.tidyverse.org/)). |
| [FunnelPlotR](https://github.com/nhs-r-community/FunnelPlotR) | An R package that extends ggplot2 for funnel plots |
| [knitr](https://github.com/yihui/knitr) | Elegant, flexible, and fast dynamic report generation with R |
| [plotly](https://plotly.com/r/) | Creates interactive and publication-ready graphs. |

The code below installs and loads the required packages into R environment.

```{r}
pacman::p_load(tidyverse, FunnelPlotR, knitr, plotly)
```

### Importing the Data

This section uses the **COVID-19_DKI_Jakarta** dataset from the [Open Data COVID-19 Provinsi DKI Jakarta portal](https://riwayat-file-covid-19-dki-jakarta-jakartagis.hub.arcgis.com/), containing cumulative cases and deaths by sub-district (e.g. kelurahan) as at 31 July 2021. The code below imports the data into R as a tibble named **covid19** and converts the City, District and sub-district variables into categorical factors.

```{r}
covid19 <- read_csv("data/COVID-19_DKI_Jakarta.csv") %>%
  mutate_if(is.character, as.factor)
```

```{r}
kable(head(covid19, n=6),format = "html")
```

## FunnelPlotR methods

The [**funnelPlotR**](https://nhs-r-community.github.io/FunnelPlotR/) package uses **ggplot2** to create funnel plots and requires a `numerator` (events of interest), a `denominator` (population at risk), and a `group` variable.

::: callout-tip
### Key arguments

-   `limit` (95 or 99 plot limits),

-   `label_outliers` (to label outlying points),

-   `Poisson_limits` and `OD_adjust` (to add Poisson or overdispersed limits),

-   `xrange` and `yrange` (to control the displayed axis ranges).

-   Other visual elements such as `title` and `label` can also be customised.
:::

The code below uses [`funnel_plot()`](https://www.rdocumentation.org/packages/rmeta/versions/3.0/topics/funnelplot) to generate a funnel plot the ratio of positive cases to deaths across sub-districts, adjusted for overdispersion. But the y-values are closely clustered and difficult to distinguish visually, so additional customisation is required to improve readability.

```{r}
funnel_plot(.data = covid19,
            numerator = Positive,
            denominator = Death,
            group = "Sub-district")
```

::: callout-tip
### Things to learn from the code chunk above:

-   In `funnel_plot()`, `group` defines the aggregation level of the points rather than a grouping aesthetic as in a scatterplot.

-   The `group` argument can be set to sub-district, district, or city, which determines how many points are plotted.

-   If `City` is selected, only six data points are shown in the funnel plot.

-   By default, the `data_type` argument is set to `"SR"`.

-   The `limit` argument controls the funnel limits, where `95` and `99` correspond to the 95% and 99.8% control limits, respectively.
:::

### FunnelPlotR methods: Makeover 1

Here we set adjust x-axis and y-axis range to display more details.

```{r}
funnel_plot(
  .data = covid19,
  numerator = Death,
  denominator = Positive,
  group = `Sub-district`,
  data_type = "PR",     #<< specifies that the plot uses a proportion or rate
  x_range = c(0, 6500),  #<< sets the x-axis range to focus on sub-districts
  y_range = c(0, 0.05),   #<< sets the y-axis range to display rate 
  theme = funnel_clean())
```

::: callout-tip
#### Things to learn from the code chunk above:

-   The `data_type` argument is used to change the default setting from `"SR"`(observed/expected, centered around 1) to `"PR"`(observed A/observed B), so the funnel plot displays proportions instead of standardised ratios,that means changing from **relative performance** to **absolute proportion**.

-   The `xrange` and `yrange` arguments are used to control the displayed ranges of the x-axis and y-axis, improving readability when values are tightly clustered.
:::

### FunnelPlotR methods: Makeover 2

Here we add chart title, axis labels and disable the labels for sub-district.

```{r}
#| code-fold: true
#| code-summary: SHOW
funnel_plot(
  .data = covid19,
  numerator = Death,
  denominator = Positive,
  group = `Sub-district`,
  data_type = "PR",   
  x_range = c(0, 6500),  
  y_range = c(0, 0.05),
  label = NA,
  title = "Cumulative COVID-19 Fatality Rate by 
  \nCumulative Total Number of COVID-19 Positive Cases", #<< sets the main title     
  x_label = "Cumulative COVID-19 Positive Cases", #<< customises the x-axis label
  y_label = "Cumulative Fatality Rate"  #<<customises the y-axis label
)

```

::: callout-tip
#### Things to learn from the code chunk above

-   `label = NA` turns off point labelling in the funnel plot to improve visual clarity.

-   **Backticks** are used to reference column names that contain nonstandard characters, such as hyphens (`Sub-district`), spaces, start with numbers etc which would otherwise be interpreted as operators in R.

-   Although`funnel_plot()` returns a ggplot object, it is not designed for **plotly** conversion, and interactive behaviour is limited or unreliable when using `ggplotly()`.
:::

## Funnel Plot for Fair Visual Comparison: ggplot2 methods

In this section, we learn hands on experience in building funnel plots step by step using **ggplot2**. The objective is to strengthen the proficiency with ggplot2 for customising specialised data visualisations such as funnel plots.

### Computing the basic derived fields

First, we manually derive the cumulative death rate and the standard error of the cumulative death rate. The formulas are as followed.

$$
\hat{p} = \frac{\text{Deaths}}{\text{Positive}}
$$ $$
SE(\hat{p}) = \sqrt{\frac{\hat{p}(1 - \hat{p})}{n}}
$$

```{r}
df <- covid19 %>%
  mutate(rate = Death/Positive)%>%
  mutate(rate.se = sqrt((rate*(1-rate))/(Positive)))%>%
  filter(rate >0) #filter out the record where rate ==0

```





```{r}
kable(head(df, n= 6),"html")
```






Next, we use`weighted.mean(x, wt)` function from **base R** to calculate the weighted mean. Weighted mean is based on inverse-variance weights so that more precise rate estimates contribute more strongly to the pooled reference value.

$$
\hat{r}_{\text{weighted}}
=
\frac{\sum_{i=1}^{n} w_i r_i}{\sum_{i=1}^{n} w_i}
=
\frac{\sum_{i=1}^{n} \frac{r_i}{SE_i^2}}{\sum_{i=1}^{n} \frac{1}{SE_i^2}}
$$

```{r}
fit.mean <- weighted.mean(df$rate, 1 / df$rate.se^2)
```

The weighted mean = 0.0150 is a single pooled reference value used as the centre of the funnel plot.

```{r}
fit.mean
```

::: callout-tip
-   `df$rate` are the observed death rates for each sub-district

-   `df$rate.se` are their standard errors

-   The weights `1 / df$rate.se^2` are inverse variances

-   Observations with smaller uncertainty (smaller SE) get more weight
:::

### Calculate lower and upper limits for 95% and 99.9% CI

We use `seq(from, to, by, ...)` to generate regular sequences of numbers from 1 to 6231, as denominator.

```{r}
max(df$Positive)
```

```{r}

summary(seq(1, max(df$Positive), 1))
```

The code chunk below is used to compute the lower and upper limits for 95% and 95% confidence interval and save the results into a table.

```{r}
number.seq <- seq(1, max(df$Positive), 1)
number.ll95 <- fit.mean - 1.96 * sqrt((fit.mean*(1-fit.mean)) / (number.seq)) 
number.ul95 <- fit.mean + 1.96 * sqrt((fit.mean*(1-fit.mean)) / (number.seq)) 
number.ll999 <- fit.mean - 3.29 * sqrt((fit.mean*(1-fit.mean)) / (number.seq)) 
number.ul999 <- fit.mean + 3.29 * sqrt((fit.mean*(1-fit.mean)) / (number.seq)) 
dfCI <- data.frame(number.ll95, number.ul95, number.ll999, 
                   number.ul999, number.seq, fit.mean)
```

```{r}
kable(head(dfCI), "html") %>%
  kableExtra::kable_styling(
    bootstrap_options = c("striped", "hover", "condensed"),
    full_width = FALSE,
    position = "center")
```

### Plotting a static funnel plot

In the code chunk below, ggplot2 functions are used to plot a static funnel plot.

::: panel-tabset
#### Plot

```{r}
#| echo: false

p <- ggplot(df, aes(x = Positive, y = rate)) +
  geom_point(
    aes(text = paste(
        "Positive:", Positive,
        "\nRate:", round(rate, 4),
        "\nSub-district:", `Sub-district`)),
    alpha = 0.8, colour = "#1877F2", size = 0.8)+
  
    # OR geom_point(aes(label=`Sub-district`),  alpha=0.4) +

  # ggrepel doesnt work properly with plotly
  #ggrepel::geom_text_repel(aes(label = `Sub-district`), max.overlaps = 2, size = 3)+
  
  #reference CI
  geom_line(data = dfCI, 
            aes(x = number.seq, y = number.ll95, linetype ="95% CI"),
            size = 0.4, colour="purple") +
  geom_line(data = dfCI, 
            aes(x = number.seq, y = number.ul95, linetype ="95% CI"),
            size = 0.4, colour="purple") +
    geom_line(data = dfCI, 
            aes(x = number.seq, y = number.ll999, linetype ="99% CI"),
            size = 0.6, colour="darkgreen", alpha = 0.4) + 
      geom_line(data = dfCI, 
            aes(x = number.seq, y = number.ul999,linetype ="99% CI"),
            size = 0.6, colour="darkgreen", alpha = 0.4) +
  # weighted mean in the center
   geom_hline(data = dfCI, 
             aes(yintercept = fit.mean, linetype = "Mean"), 
             size = 0.5, 
             colour = "grey40") +
  # up until now the plot is focused around 0, zoom in
  coord_cartesian(ylim=c(0,0.05)) + 
  scale_linetype_manual(
    name = "Reference",
    values = c("95% CI" = "longdash", "99% CI" = "solid", "Mean" = "dashed")) +
  annotate("text", x = 1, y = -0.13, label = "95%", size = 3, colour = "grey40") + 
  annotate("text", x = 4.5, y = -0.18, label = "99%", size = 3, colour = "grey40") + 
  ggtitle("Cumulative Fatality Rate by Cumulative Number of COVID-19 Cases") +
  xlab("Cumulative Number of COVID-19 Cases") + 
  ylab("Cumulative Fatality Rate") +
  theme_light() +
  theme(plot.title = element_text(size=12), # control title text size
        legend.position = c(0.91,0.85),     # place legend inside plot area
        legend.title = element_text(size=7),  # reduce legend title size
        legend.text = element_text(size=7),   # reduce legend text size
        legend.background = element_rect(colour = "grey60", linetype = "dotted"),  # legend box
        legend.key.height = unit(0.3, "cm"))   # reduce legend spacing
p
```

#### Code

```{r}
#| eval: false

p <- ggplot(df, aes(x = Positive, y = rate)) +
  geom_point(
    aes(text = paste(
        "Positive:", Positive,
        "\nRate:", round(rate, 4),
        "\nSub-district:", `Sub-district`)),
    alpha = 0.8, colour = "#1877F2", size = 0.8)+
  
    # OR geom_point(aes(label=`Sub-district`),  alpha=0.4) +

  # ggrepel doesnt work properly with plotly
  #ggrepel::geom_text_repel(aes(label = `Sub-district`), max.overlaps = 2, size = 3)+
  
  #reference CI
  geom_line(data = dfCI, 
            aes(x = number.seq, y = number.ll95, linetype ="95% CI"),
            size = 0.4, colour="purple") +
  geom_line(data = dfCI, 
            aes(x = number.seq, y = number.ul95, linetype ="95% CI"),
            size = 0.4, colour="purple") +
    geom_line(data = dfCI, 
            aes(x = number.seq, y = number.ll999, linetype ="99% CI"),
            size = 0.6, colour="darkgreen", alpha = 0.4) + 
      geom_line(data = dfCI, 
            aes(x = number.seq, y = number.ul999,linetype ="99% CI"),
            size = 0.6, colour="darkgreen", alpha = 0.4) +
  # weighted mean in the center
   geom_hline(data = dfCI, 
             aes(yintercept = fit.mean, linetype = "Mean"), 
             size = 0.5, 
             colour = "grey40") +
  # up until now the plot is focused around 0, zoom in
  coord_cartesian(ylim=c(0,0.05)) + 
  scale_linetype_manual(
    name = "Reference",
    values = c("95% CI" = "longdash", "99% CI" = "solid", "Mean" = "dashed")) +
  annotate("text", x = 1, y = -0.13, label = "95%", size = 3, colour = "grey40") + 
  annotate("text", x = 4.5, y = -0.18, label = "99%", size = 3, colour = "grey40") + 
  ggtitle("Cumulative Fatality Rate by Cumulative Number of COVID-19 Cases") +
  xlab("Cumulative Number of COVID-19 Cases") + 
  ylab("Cumulative Fatality Rate") +
  theme_light() +
  theme(plot.title = element_text(size=12), # control title text size
        legend.position = c(0.91,0.85),     # place legend inside plot area
        legend.title = element_text(size=7),  # reduce legend title size
        legend.text = element_text(size=7),   # reduce legend text size
        legend.background = element_rect(colour = "grey60", linetype = "dotted"),  # legend box
        legend.key.height = unit(0.3, "cm"))   # reduce legend spacing
p
```

:::

::: callout-tip
#### Line-type

-   0 ("blank"): No line.

-   1 ("solid"):the default

-   2 ("dashed"): A line made of short, equal-length segments with gaps.

-   3 ("dotted"): A line composed of dots.

-   4 ("dotdash"): A pattern of a dot followed by a dash.

-   5 ("longdash"): A line made of longer dashes.

-   6 ("twodash"): A pattern of two short dashes followed by a gap.

-   **Legends only appear when we map aesthetics (colour, shape, linetype, etc.) to variables within aes().**
:::

### Interactive Funnel Plot: plotly + ggplot2

The interactive funnel plot could be created by using `ggplotly()` of **plotly** package.

```{r}
fp_ggplotly <- ggplotly(p,tooltip = "text")
# OR  tooltip = c("label", "x", "y"))
fp_ggplotly
```

::: callout-caution
-   `text` is an aesthetic that belongs to **plotly**, so it is ignored when the plot is first rendered using **static ggplot2**.

-   **ggrepel** does not work reliably with **plotly**.
:::


::: callout-note
-  Early in an outbreak or in a small populations, fatality rates look **wide**. As more case accumulated, fatality rats settled down and **converaged** toward a stable average.

- The CI bands were **wide** at low cumulative case counts due to high sampling uncertainty and progressively **narrowed** as case number increase. This implies more stable fatality rate estimates.

:::






## Reference

Kam, Tin Seong (2026). [8 Funnel Plots for Fair Comparisons](https://r4va.netlify.app/chap08) R for Visual Analytics.
