---
title: "06: A Basic Shiny App"
author: "Yiqiong PAN"
date: "21 Feb, 2026"
date-modified: last-modified

execute:
  echo: true    # display the code
  eval: true    # run the code
  message: false
  warning: false
  freeze: true  # do not re-render if nothing edited

format: 
  html:
    toc: true
    code-link: true

number-sections: true
number-depth: 6

editor: visual
---

## Learning Outcome

By the end of this hands-on exercise, we will be able to create a basic Shiny application using R packages:

## Getting Started

There are **2** packages that will be used in this exercise.

| Package | Description |
|------------------------------------|------------------------------------|
| [tidyverse](https://www.tidyverse.org/) | A collection of R packages for data importing, cleaning, transformation, and visualisation (e.g., [readr](https://readr.tidyverse.org/), dplyr, tidyr, [ggplot2](https://ggplot2.tidyverse.org/)). |
| [shiny](https://shiny.posit.co/) | Builds interactive web-based applications for data science |

## app.R structure in Shiny

Shiny apps are commonly organised in a single file called app.R. A basic app has three main parts: packages, UI, and server, then you run the app.

### Packages

Load the packages we need at the top of the file using `library()`. We should check is whether a package provides certain UI functions (like `plotOutput()`, `DTOutput()`) or server rendering functions (like `renderPlot()`, `renderDT()`), depending on what we want to display.

### UI

The UI describes the layout and inputs the user can interact with. A common pattern is:

```{r}
#| eval: false
ui <- fluidPage(...)
```

`fluidPage()` adjusts to different screen sizes (desktop, mobile, etc).

**Common fluidPage() components** include:

-   **titlePanel()** creates the main page title.

-   **Layout** panels can be created in different ways.

A classic layout is:

```{r}
#| eval: false
sidebarLayout(sidebarPanel(...), 
                mainPanel(...))
```

::: callout-tip
#### Notes

-   sidebarPanel() holds **inputs** (controls).

-   mainPanel() holds **outputs** (plots, tables, text, etc).
:::

**Inputs inside sidebarPanel()**

Common inputs include:

-   **selectInput**(inputId = "...", label, choices, selected = NULL, multiple = FALSE, selectize = TRUE, width = NULL, size = NULL)

-   **sliderInput**(inputId = "...", label, min, max, value, step = NULL, round = FALSE, ticks = TRUE, animate = FALSE, width = NULL, sep = ",", pre = NULL, post = NULL, timeFormat = NULL, timezone = NULL, dragRange = TRUE)

::: callout-tip
#### Notes

-   In Shiny, each page row is divided into **12** columns total. By default: sidebarPanel is 4-column wide and mainPanel is 8-column wide. We can change the widths,e.g. setting the `width` 3 in `sidbarPanel()` and 9 in `mainPanel()`.

-   Internally, Shiny is doing **fluidRow**(column(width = x, "location"),...), so the width is adjusted automatically for different screen sizes. **Please note**: he sum of width should not exceed 12, if less than 12, Shiny leaves empty space, if greater than 12, the layout looks odd.

-   Good practice during storyboarding is to plan **unique** inputId values early (for example inputId = "A", "B", "C" or meaningful names) to avoid duplicates, because duplicated IDs cause **bugs**.
:::

### Server

The server defines the logic and computation. It reacts to user inputs and generates outputs.

Typical structure:

```{r}
#| eval: false
server <- function(input, output, session) { ... }
```

Inside server, `output$...` is a Shiny server object we create.

For Example:

```{r}
#| eval: false
output$plot1 <- renderPlot({ ... })
output$table1 <- renderTable({ ... })
```

### Run the app

Finally `shinyApp()` tells Shiny to use the UI and server and launch an interactive web service.

```{r}
#| eval: false
shinyApp(ui = ui, server = server)
```

### Deleting a deployed Shiny app

On *shinyapps.io* platform, the usual safe approach is to **archive** first, then **delete**.

### Example - prototype1

[Here](https://ypan608.shinyapps.io/Prototype1/) is the prototype we tried out in the class.

```{r}
#| eval: false
pacman::p_load(shiny, tidyverse)


exam <- read_csv("data/Exam_data.csv")

#print(exam)

# Define UI for application that draws a histogram
ui <- fluidPage(
  titlePanel("Pupils Exam Results Dashboard"),
  sidebarLayout(
    sidebarPanel(
      selectInput(inputId = "variable",
                  label = "Subject:",
                  choices = c("English" = "ENGLISH",
                              "Maths" = "MATHS",
                              "Science" = "SCIENCE"),
                  selected = "ENGLISH"), # default choice
      sliderInput(inputId = "bins",
                  label = "Number of Bins",
                  min = 5,
                  max = 20,
                  value = 10)
    ),
    mainPanel(
      plotOutput("distPlot")
    )
  )
)

  
# Define server logic required to draw a histogram
server <- function(input, output) {
  output$distPlot <- renderPlot({
    ggplot(data = exam, 
           aes_string(x = input$variable)) +
             geom_histogram(bins = input$bins,
                            color = "black",
                            fill = "lightblue")
  })
}

# Run the application 
shinyApp(ui = ui, server = server)
```

## Reference

Kam, Tin Seong (2026). [Designing Web-enabled Visual Analytics Application with Shiny](https://isss608-ay2025-26jan.netlify.app/outline/lesson06_outline), ISSS608 Visual Analytics and Applications
