---
title: "10 Information Dashboard Design: R methods"
author: "Yiqiong PAN"
date: "Mar 15, 2026"
date-modified: "last-modified"
execute:
  echo: true    # display the code
  eval: true    # run the code
  message: false
  warning: false
  freeze: true  # do not re-render if nothing edited

format: 
  html:
    toc: true
    code-link: true

number-sections: true
number-depth: 6

editor: visual
---

## Overview

In this hands on exercise, we learn how to build information dashboard in R.

By the end of this exercise, we will be able to:

-  create bullet charts using **ggplot2**

-  create sparklines using **ggplot2**

- build an industry standard dashboard using **R Shiny**


## Getting Started

### Installing and Loading Packages

In this hands on exercise, four R packages will be used. 


| Package | Description |
|------------------------------------|------------------------------------|
| [tidyverse](https://www.tidyverse.org/) | A collection of R packages for data importing, cleaning, transformation, and visualisation (e.g., [readr](https://readr.tidyverse.org/), [dplyr](https://dplyr.tidyverse.org/), [tidyr](https://tidyr.tidyverse.org/), [ggplot2](https://ggplot2.tidyverse.org/)), [tibble](https://tibble.tidyverse.org/),[stringr](https://stringr.tidyverse.org/),[forcats](https://forcats.tidyverse.org/), and [purr](https://purrr.tidyverse.org/)|
| [gtExtras](https://jthomasmock.github.io/gtExtras/index.html) | Provides additional helper functions that make it easier to create visually appealing tables with gt. The [gt](https://gt.rstudio.com/) package is designed to help users produce well formatted and attractive tables in R. |
| [ggthemes](https://cran.r-project.org/web/packages/ggthemes/index.html) | Provides 'ggplot2' themes and scales that replicate some of the famous looks. |
| [lubridate](https://cran.r-project.org/web/packages/lubridate/index.html) | A Tidyverse package that handles date-time objects |
| [reactable](https://glin.github.io/reactable/index.html) | Provides functions for creating interactive data tables in R, built on the [React Table](https://github.com/tanstack/table/tree/v7) library and implemented through [reactR](https://github.com/react-R/reactR).|
| [reactablefmtr](https://kcuilla.github.io/reactablefmtr/index.html) | Offers a range of features that simplify and improve the styling of interactive reactable tables, providing easy to use and highly customisable functions and themes.
|

```{r}
pacman::p_load(lubridate, ggthemes, reactable,
reactablefmtr, gt, gtExtras, tidyverse)
```

## Importing Microsoft Access Database

### The data set

For the purpose of this study, a personal database in Microsoft Access *mdb* format called **Coffee Chain** will be used.

### Importing database into R

In the code chunk below, `odbcConnectAccess()` from the [RODBC](https://cran.r-project.org/web/packages/RODBC/index.html) package is used to import a database query table into **R**.


```{r}
library(RODBC)
con <- odbcConnectAccess2007('data/Coffee Chain.mdb')
coffeechain <- sqlFetch(con, 'CoffeeChain Query')
write_rds(coffeechain, "data/CoffeeChain.rds")
odbcClose(con)
```


 An **.mdb** file is a Microsoft Access database file that stores tables, queries, and other database objects.

Next, `sqlFetch()` is used to retrieve the table named *CoffeeChain Query* from the database. The retrieved data is returned to R as a data frame.

Finally, `write_rds()` is used to save the data frame as an **.rds file**. This format stores a single R object efficiently and allows the dataset to be quickly reloaded later using `read_rds()` for further analysis.


```{r}
knitr::kable(head(coffeechain, 10))
```

:::callout-tip

- Before running the code chunk, the R environment should be set to the 32 bit version because `odbcConnectAccess()` relies on 32 bit ODBC drivers and may not work properly in a 64 bit R session. The system architecture can be verified using `sessionInfo()` or `.Machine$sizeof.pointer`.

- I am using R 4.5.2, and the platform shown in `sessionInfo()` is x86_64-w64-mingw32/x64, which indicates a 64 bit R system. In the installation directory, there is no `i386` folder in the bin directory, only the 64 bit version of R is installed. Despite this, the database query table was imported successfully.

:::


### Data Preparation

The code chunk below is used to import *CoffeeChain.rds* into R.

```{r}
coffeechain <- read_rds("data/CoffeeChain.rds")
```

The code below aggregates Sales as `current` and aggregates Budgeted Sales as `target`  by Product by using `group_by()`, `summarise()` and `ungroup()` from **dlpyr** package.

```{r}
product <- coffeechain %>%
  group_by(`Product`) %>%
  summarise(`target` = sum(`Budget Sales`),
            `current` = sum(`Sales`)) %>%
  ungroup()         
```

### Bullet chart in ggplot2


There are two types of bar charts in **ggplot2**: `geom_bar()` and [`geom_col()`](https://ggplot2.tidyverse.org/reference/geom_bar.html).

**`geom_bar()`** sets the height of each bar based on the number of observations in each category. If a **weight** aesthetic is provided, the height represents the sum of the weights.

In contrast, **`geom_col()`** uses the actual values from the dataset to determine the height of the bars. Therefore, it is appropriate when the values to be displayed are already computed.

We finally have a good example of `geom_col()` here. The dataset has the prepared **current** and **target** values for each product. The code belpw creates a bar plot that compares the current performance with the target. Different background shades and bar widths are used to represent the target ranges and the current value, forming a cool **bullet chart** style visualisation. The wider grey bars indicate the target thresholds, while the narrower black bar represents the current value.

[`geom_errorbar()`](https://ggplot2.tidyverse.org/reference/geom_linerange.html) is cleverly designed here to draw a vertical red line as target value for each product. Because the minimum and maximum values are set to the same value (target), the error bar collapses into a line that serves as a reference marker rather than showing variability.

Now we can easily pinpoint which products meet their targets and which do not.

```{r}
ggplot(product, aes(Product, current)) +
  geom_col(aes(Product, max(target) * 1.01),
           fill = "grey85", width = 0.85) +
  geom_col(aes(Product, target * 0.75),
           fill = "grey60", width = 0.85) +
  geom_col(aes(Product, target * 0.5),
           fill = "grey50", width = 0.85) +
  geom_col(aes(Product, current), width = 0.35, fill = "black")+
   geom_errorbar(aes(y = target,x = Product, min = target, ymax= target), 
                width = .4,colour = "red",linewidth = 1) +
  scale_y_continuous(labels = scales::comma) +
  coord_flip() +
  labs(x = "Product", y = "Current Sales", title  = "Coffee Chain Product Sales Dashboard") +
 theme(axis.ticks.y = element_blank()) 
 
```
:::callout-tip

In ggplot2, `coord_flip()` only changes the visual orientation (such as axis ticks, labs), but the scales still refer to the original mapping.


:::


## Plotting sparklines using ggplot2

In this section, we learn how to plot sparklines by using **ggplot2**.


### Preparing the data



## Reference

Kam, Tin Seong (2026). 31  Information Dashboard Design: R methods https://r4va.netlify.app/chap31
