---
title: "08-2 Visualising Geospatial Point Data"
author: "Yiqiong PAN"
date: "Mar 02, 2026"
date-modified: "last-modified"
execute:
  echo: true    # display the code
  eval: true    # run the code
  message: false
  warning: false
  freeze: true  # do not re-render if nothing edited

format: 
  html:
    toc: true
    code-link: true

number-sections: true
number-depth: 6

editor: visual
---

## Overview

Proportional symbol maps, also called graduated symbol maps, use symbol size to represent the magnitude of a variable, such as counts of populations. They can be either classed (graduated symbols) or unclassed (proportional symbols), where the symbol area is directly proportional to the data value.

In this exercise, we will use the **tmap** package in R to create a proportional symbol map showing the number of wins at Singapore Pools outlets.

### Learning Outcome

By the end of this hands on exercise, we will be able to use appropriate R packages to:

- Import an aspatial data file into R.
- Convert it into an `sf` point object and assign the correct coordinate reference system.
- Create interactive proportional symbol maps.



## Getting tarted

There are **3** packages that will be used in this exercise.

| Package | Description |
|------------------------------------|------------------------------------|
| [tidyverse](https://www.tidyverse.org/) | A collection of R packages for data importing, cleaning, transformation, and visualisation (e.g., [readr](https://readr.tidyverse.org/), [dplyr](https://dplyr.tidyverse.org/), [tidyr](https://tidyr.tidyverse.org/), [ggplot2](https://ggplot2.tidyverse.org/)). |
| [tmap](https://r-tmap.github.io/tmap/) | Creates cartographic quality static and interactive thematic (choropleth) maps. |
| [sf](https://r-spatial.github.io/sf/) | Handles importing, managing, and processing vector-based geospatial data |

```{r}
pacman::p_load(sf,tmap,tidyverse)
```

## Geospatial Data Wrangling

### The Data

The dataset used in this hands on exercise is **SGPools_svy21**, provided in CSV format.

The figure below shows the first 15 records of `SGPools_svy21.csv`. The file contains seven columns. The `XCOORD` and `YCOORD` fields store the projected x and y coordinates of Singapore Pools outlets and branches, using the [**SVY21** coordinate system](https://app.sla.gov.sg/sirent/About/PlaneCoordinateSystem).


### Data Import and Preparation

we use [**`read_csv()`**](https://www.rdocumentation.org/packages/qtl2/versions/0.32/topics/read_csv) to import the csv file as a tibble dataframe and name it *sgpools*.

```{r}
sgpools <- read_csv("data/aspatial/SGPools_svy21.csv")
list(sgpools)
```

### Creating a sf data.frame from an Aspatial data.frame

The code chunk below converts the `sgpools` data frame into a simple features data frame using the [`st_as_sf()`](https://r-spatial.github.io/sf/reference/st_as_sf.html) function from the **sf** package.

```{r}
sgpools_sf <- st_as_sf(sgpools, 
                       coords = c("XCOORD", "YCOORD"),
                       crs= 3414)
```

:::callout-tip


- The `coords` argument requires the x coordinate column name first, followed by the y coordinate column name.

- The `crs` argument specifies the coordinate reference system in EPSG format. [**EPSG:3414**](https://epsg.io/3414) corresponds to the Singapore SVY21 projected coordinate system. You can look up other EPSG codes at [epsg.io](https://epsg.io/).

:::

When we print the sf object, the console returns the first 10 rows and basic information of the object, such as it contains 306 point features.

```{r}
sgpools_sf
```

We can also use `class()`,`st_crs()` to verify the class and Coordinate Reference System of the sgpools_sf file.
```{r}
class(sgpools_sf)
st_crs(sgpools_sf)
```



## Drawing Proportional Symbol Map

To create an interactive proportional symbol map in R, the **view** mode of **tmap** is used.
The code chunk below switches **tmap** to interactive mode.

```{r}
tmap_mode("view")
```


### It all started with an interactive point symbol map

The code chunks below are used to create an interactive point symbol map.

```{r}
tm_shape(sgpools_sf) + 
  tm_bubbles(fill = "red",
           size = 1,
           col = "black",
           lwd = 1)
```


### Let's make it proportional

To draw a proportional symbol map, a numeric variable must be mapped to the **size** visual attribute. The code chunk below assigns the variable `Gp1Gp2Winnings` to the size aesthetic.


```{r}
tm_shape(sgpools_sf) + 
  tm_bubbles(fill = "red",
             size = "Gp1Gp2 Winnings",
             col = "black",
             lwd = 1)
```


### Let's give it a different colour

The proportional symbol map can be enhanced by adding a colour visual attribute. In the code chunk below, the `OUTLET_TYPE` variable is mapped to the colour aesthetic.


```{r}
tm_shape(sgpools_sf) + 
  tm_bubbles(fill = "OUTLET TYPE", 
             size = "Gp1Gp2 Winnings",
             col = "black",
             lwd = 1)
```


###  I have a twin brother :)

An impressive and little known feature of **tmap** view mode is that it also supports faceted maps. The `sync` argument in `tm_facets()` can be used to display multiple maps with synchronised zoom and pan behaviour.

```{r}
tm_shape(sgpools_sf) + 
  tm_bubbles(fill = "OUTLET TYPE", 
             size = "Gp1Gp2 Winnings",
             col = "black",
             lwd = 1) + 
  tm_facets(by= "OUTLET TYPE",
            nrow = 1,
            sync = TRUE)
```


```{r}
tmap_mode("plot")
```


## Reference


Kam, Tin Seong (2026). 22 Visualising Geospatial Point Data, https://r4va.netlify.app/chap22

