---
title: "02 Beyond ggplot2 Fundamentals"
author: "Yiqiong PAN"
date: "Jan 19, 2026"
date-modified: last-modified

execute:
  echo: true    # display the code
  eval: true    # run the code
  message: false
  warning: false
  freeze: true  # do not re-render if nothing edited

format: 
  html:
    toc: true
    code-link: true

number-sections: true
number-depth: 6

editor: visual
---

## Overview

In this chapter, we will learn ggplot2 extensions for creating more professional statistical graphics. By the end of this exercise, we will be able to:

-   improve annotation placement using **ggrepel**

-   apply publication-ready themes with **ggthemes** and **hrbrthemes**

-   combine multiple plots using **patchwork**

## Getting Started

### Installing and Loading the Required Libraries

There are five packages that will be used in this exercise.

| Package | Description |
|------------------------------------|------------------------------------|
| [tidyverse](https://www.tidyverse.org/) | A collection of R packages for data importing, cleaning, transformation, and visualisation (e.g., [readr](https://readr.tidyverse.org/), dplyr, tidyr, [ggplot2](https://ggplot2.tidyverse.org/)). |
| [ggrepel](https://cran.r-project.org/web/packages/ggrepel/vignettes/ggrepel.html) | Provides geoms for non-overlapping text labels in ggplot2 |
| [ggthemes](https://cran.r-project.org/web/packages/ggthemes/refman/ggthemes.html) | Provides **ggplot2** themes, geoms and scales that replicate the famous styles ( e.g. Tufte, Economist, Tufte) |
| [hrbrthemes](https://github.com/hrbrmstr/hrbrthemes) | Provides minimal and professional looking themes and better typography and styleing |
| [patchwork](https://cran.r-project.org/web/packages/patchwork/refman/patchwork.html) | Extends ggplot2 by letting user combine plots easily using operators. |

The code below installs and loads the required packages into R environment. **hrbrthemes** package can be installed using the **remote** package, the code can be found on its github [ReadMe](https://github.com/hrbrmstr/hrbrthemes).

```{r}
pacman::p_load(tidyverse, ggrepel, ggthemes, hrbrthemes, patchwork)
```

The code chunk below solves the system warning when using **hrbthemes** on Windows.

```{r}
#| code-summary: Font Setup
extrafont::loadfonts(device = "win")
sysfonts::font_add_google("Roboto Condensed", "Roboto Condensed")
showtext::showtext_auto()
```


### Importing the Data

A CSV file containing the final exam results of 322 local Primary 3 students has been imported into the R environment by using function [`read_csv()`](https://readr.tidyverse.org/reference/read_delim.html) from [**readr**](https://readr.tidyverse.org/) package, one of tidyverse family. The dataset includes variables such as student ID, class, gender, race, and grades for English, Maths, and Science.

```{r}
exam_data <- read_csv("data/Exam_data.csv")
```

## Beyond ggplot2 Annotation: ggrepel

To visualise the Maths and English scores for each student in a scatterplot, we can use `geom_point()` and `geom_smooth()` to add a reference line. However since all the student IDs are shown on the chart, the labels overlap and become difficult to read.

:::{.panel-tabset}
### The plot

```{r}
#| echo: false
ggplot(data = exam_data,
       aes(x= MATHS, y = ENGLISH)) +
         geom_point() +
    geom_smooth(method = lm, linewidth = 0.5) +
  geom_label(aes(label = ID, hjust = 0.5, vjust= -0.5)) +
  coord_cartesian(xlim = c(0,100), ylim = c(0,100)) +
  ggtitle("English scores VS Math scores for Primary 3")
```

### The code

```{r}
#| eval: false
ggplot(data = exam_data,
       aes(x= MATHS, y = ENGLISH)) +
         geom_point() +
    geom_smooth(method = lm, linewidth  = 0.5) +
  geom_label(aes(label = ID, hjust = 0.5, vjust= -0.5)) + 
  coord_cartesian(xlim = c(0,100), ylim = c(0,100)) +
  ggtitle("English scores VS Math scores for Primary 3")
```
:::

Here we could use functions like [`geom_text_repel()`](https://ggrepel.slowkow.com/reference/geom_text_repel) or [`geom_label_repel()`](https://ggrepel.slowkow.com/reference/geom_text_repel#geom-label-repel) from **ggrepel** package to customise the label geoms for **ggplot2**. The difference is that `geom_label_repel()` adds a rectangle behind the text, which helps the labels stand out from the background, similar to `geom_label()` vs `geom_text()` that it shows plain text only.

### Working with ggrepel

ggrepel automatically places labels nicely to minimise overlap. However, we can still force all labels to be displayed by using the argument `max.overlaps = Inf`. The text font can be adjusted using the `family` argument, while the font style can be changed using `fontface.` In addition,`box.padding` controls the spacing between labels, the smaller the value, the closer the labels will be.

:::{.panel-tabset}
#### The plot - label

```{r}
#| echo: false
ggplot(data = exam_data,
       aes(x= MATHS, y = ENGLISH)) +
         geom_point() +
    geom_smooth(method = lm, linewidth  = 0.5) +
  geom_label_repel(aes(label = ID, fontface = "bold")) +
  coord_cartesian(xlim = c(0,100), ylim = c(0,100)) +
  ggtitle("English scores VS Math scores for Primary 3")
```

#### The code - label

```{r}
#| eval: false
ggplot(data = exam_data,
       aes(x= MATHS, y = ENGLISH)) +
         geom_point() +
    geom_smooth(method = lm, linewidth  = 0.5) +
  geom_label_repel(aes(label = ID, fontface = "bold")) + 
  coord_cartesian(xlim = c(0,100), ylim = c(0,100)) +
  ggtitle("English scores VS Math scores for Primary 3")
```

#### The plot - text

```{r}
#| echo: false
ggplot(data = exam_data,
       aes(x= MATHS, y = ENGLISH)) +
         geom_point() +
    geom_smooth(method = lm, linewidth  = 0.5) +
   geom_text_repel(aes(label = ID), 
                  family = "serif", box.padding = 0.4)+
  coord_cartesian(xlim = c(0,100), ylim = c(0,100)) +
  ggtitle("English scores VS Math scores for Primary 3")
```

#### The code - text

```{r}
#| eval: false
ggplot(data = exam_data,
       aes(x= MATHS, y = ENGLISH)) +
         geom_point() +
    geom_smooth(method = lm, linewidth  = 0.5) +
   geom_text_repel(aes(label = ID), 
                  family = "serif", box.padding = 0.5)+
  coord_cartesian(xlim = c(0,100), ylim = c(0,100)) +
  ggtitle("English scores VS Math scores for Primary 3")
```
:::

## Beyond ggplot2 Themes: ggthemes

There are 8 [built-in themes](https://ggplot2.tidyverse.org/reference/ggtheme.html) in ggplot2. They are: `theme_gray()`(default), `theme_bw()`, `theme_classic()`, `theme_dark()`, `theme_light()`, `theme_linedraw()`, `theme_minimal()`, and `theme_void()`, some of them we tried out in [Hand-on Exercise 1.11](https://isss608-ypan.netlify.app/hands-on_ex/hands-on_ex01/hands-on_ex01#essential-grammatical-elements-in-ggplot2-themes).

:::{.panel-tabset}
### The plot - linedraw

```{r}
#| code-fold: true
#| code-summary: "VIEW CODE"
ggplot(data = exam_data,
       aes(x= MATHS)) +
         geom_histogram(bins = 20, boundary = 100, fill = "grey10", colour = "grey25") +
  ggtitle("Maths scores for Primary 3") +
  theme_linedraw()

```

### The plot - dark

```{r}
#| code-fold: true
#| code-summary: "VIEW CODE"
ggplot(data = exam_data,
       aes(x= MATHS)) +
         geom_histogram(bins = 20, boundary = 100, fill = "grey10", colour = "grey25") +
  ggtitle("Maths scores for Primary 3") +
  theme_dark()

```

### The plot - blackwhite

```{r}
#| code-fold: true
#| code-summary: "VIEW CODE"
ggplot(data = exam_data,
       aes(x= MATHS)) +
         geom_histogram(bins = 20, boundary = 100, fill = "grey10", colour = "grey25") +
  ggtitle("Maths scores for Primary 3") +
  theme_bw()

```

### The plot - void

```{r}
#| code-fold: true
#| code-summary: "VIEW CODE"
ggplot(data = exam_data,
       aes(x= MATHS)) +
         geom_histogram(bins = 20, boundary = 100, fill = "lightblue", colour = "grey25") +
  ggtitle("Maths scores for Primary 3") +
  theme_void()

```
:::

### Working with ggtheme Package

[**ggthemes**](https://cran.r-project.org/web/packages/ggthemes/index.html) mimics [famous plotting styles](https://yutannihilation.github.io/allYourFigureAreBelongToUs/ggthemes/) such as The Wall Street Journal `theme_wsj()`, The Economist `theme_economist()`, Stata `theme_stata()`, Excel `theme_excel()`, Edward Tufte `theme_tufte()`, Stephen Few `theme_few()`, and FiveThirtyEight`theme_fivethirtyeight()`.

Here we try all out again.

:::{.panel-tabset}
#### Stata

```{r}
#| code-summary: "VIEW CODE"
#| code-fold: true
ggplot(data = exam_data,
       aes(x= MATHS)) +
         geom_histogram(bins = 20, boundary = 100, fill = "lightblue", colour = "grey25") +
  ggtitle("Maths scores for Primary 3") +
  theme_stata()

```

#### The Economist

```{r}
#| code-summary: "VIEW CODE"
#| code-fold: true
ggplot(data = exam_data,
       aes(x= MATHS)) +
         geom_histogram(bins = 20, boundary = 100, fill = "lightblue", colour = "grey25") +
  ggtitle("Maths scores for Primary 3") +
  theme_economist()
```

#### Excel

```{r}
#| code-summary: "VIEW CODE"
#| code-fold: true
ggplot(data = exam_data,
       aes(x= MATHS)) +
         geom_histogram(bins = 20, boundary = 100, fill = "lightblue", colour = "grey25") +
  ggtitle("Maths scores for Primary 3") +
  theme_excel()
```

#### The WSJ

```{r}
#| code-summary: "VIEW CODE"
#| code-fold: true
ggplot(data = exam_data,
       aes(x= MATHS)) +
         geom_histogram(bins = 20, boundary = 100, fill = "lightblue", colour = "grey25") +
  ggtitle("Maths scores for Primary 3") +
  theme_wsj()
```

#### Tufte

```{r}
#| code-summary: "VIEW CODE"
#| code-fold: true
ggplot(data = exam_data,
       aes(x= MATHS)) +
         geom_histogram(bins = 20, boundary = 100, fill = "lightblue", colour = "grey25") +
  ggtitle("Maths scores for Primary 3") +
  theme_tufte()
```

#### Few

```{r}
#| code-summary: "VIEW CODE"
#| code-fold: true
ggplot(data = exam_data,
       aes(x= MATHS)) +
         geom_histogram(bins = 20, boundary = 100, fill = "lightblue", colour = "grey25") +
  ggtitle("Maths scores for Primary 3") +
  theme_few()
```

#### FiveThirtyEight

```{r}
#| code-summary: "VIEW CODE"
#| code-fold: true
ggplot(data = exam_data,
       aes(x= MATHS)) +
         geom_histogram(bins = 20, boundary = 100, fill = "lightblue", colour = "grey25") +
  ggtitle("Maths scores for Primary 3") +
  theme_fivethirtyeight()
```
:::

### Working with hrbthemes Package

[**hrbthemes**](https://github.com/hrbrmstr/hrbrthemes) provides typographic themes and components for ggplot2 e.g. themes(`theme_ipsum()`), scales (`scale_y_percent()`), palettes(`ft_cols()`), fonts(`font_ps()`), Bit12 colour scales (`scale_fill_bit12_continuous()`), Flexoki palettes (`flexoki_light()`), utilities (`flush_ticks()`).

:::{.panel-tabset}
#### theme_ipsum()

```{r}
#| code-fold: true
#| code-summary: "VIEW CODE"
ggplot(data = exam_data, aes(x = MATHS)) +
  geom_histogram(bins = 20, boundary = 100,
                 colour = "grey25", fill = "grey90") +
  ggtitle("Distribution of Maths Scores")+
   theme_ipsum()
```

#### theme_ipsum() with Styling

The **hrbrthemes** package has two main goals: improving typography in plots (such as label placement and fonts) and supporting efficient production workflows.

A [**production workflow**](https://hrbrmstr.github.io/hrbrthemes/articles/why_hrbrthemes.html) refers to creating visuals for publication, such as blog posts, papers, reports, or presentations. While rough plots are useful during analysis, polished final graphics can help capture attention and improve audience engagement.

```{r}
#| code-fold: true
#| code-summary: "VIEW CODE"
ggplot(data = exam_data, aes(x = MATHS)) +
  geom_histogram(bins = 20, boundary = 100, color="grey25", 
                 fill="grey90") +
  labs(title = "Distribution of Maths Scores", caption = "Primary 3 students' year-end scores") +
  theme_ipsum(axis_title_size = 18,
              base_size = 15,
              grid = "Y")
```
:::

:::{.callout-tip}

#### Tips

-   `axis_title_size` controls the axis title font size

-   `base_size` Sets the default text size for the theme

-   `grid` Controls whether grid lines appear. If set to "Y", it shows grid lines mainly to help track values along the y-axis (making it easier to compare heights/levels).

-   `base_family` is used to set the font family. On Windows, the default fonts used by `theme_ipsum()` (Arial Narrow) and `theme_ft_rc()` (Roboto Condensed) may not be installed, which causes warning messages. If `base_family` is not specified, R will fall back to the default `sans` font. However, we can use `font_import()` from **extrafont** package to import the fonts from system fonts folder to a R font database (the process might take a few minutes) and then load the fonts into the R environment by using `loadfonts(device = "win")`. Alternatively, we use `font_add_google()` from **sysfonts** package to download the fonts and load the fonts using `showtext_auto()` from **showtext** package.

:::

## Beyond Single Graph

We usually need multiple graphs to tell a compelling visual story. Several ggplot2 extensions provide functions for composing figures made up of multiple plots. In this section, we learn how to create a composite plot by combining multiple graphs. First, let us create three statistical graphics using the code chunk below.

:::{.panel-tabset}
### p1

```{r}
#| code-fold: true
#| code-summary: "VIEW CODE"

p1<- ggplot(data = exam_data,
            aes(x=MATHS)) +
  geom_histogram(bins = 20, boundary = 100,
                 colour = "grey25", fill = "grey90") +
  coord_cartesian(xlim = c(0,100)) + 
  ggtitle("Distribution of Maths Scores")

p1

```

### p2

```{r}
#| code-fold: true
#| code-summary: "VIEW CODE"

p2<- ggplot(data = exam_data,
            aes(x=ENGLISH)) +
  geom_histogram(bins = 20, boundary = 100,
                 colour = "grey25", fill = "grey90") +
  coord_cartesian(xlim = c(0,100)) + 
  ggtitle("Distribution of English Scores")

p2
```

### p3

```{r}
#| code-fold: true
#| code-summary: "VIEW CODE"

p3<- ggplot(data = exam_data,
            aes(x=MATHS, y =ENGLISH)) +
  geom_point() +
  geom_smooth(method = lm, linewidth= 0.5) +
  coord_cartesian(xlim = c(0,100), ylim = c(0,100)) + 
  ggtitle("English Scores versus Maths Scores for Primary 3")

p3
```
:::

### Creating Composite Graphics: pathwork Methods

Several ggplot2 extension packages support composite figures, such as [`grid.arrange()`](https://cran.r-project.org/web/packages/gridExtra/vignettes/arrangeGrob.html) from [**gridExtra**](https://cran.r-project.org/web/packages/gridExtra/refman/gridExtra.html) package and [`plot_grid()`](https://wilkelab.org/cowplot/reference/plot_grid.html) from [**cowplot**](https://wilkelab.org/cowplot/index.html). In this section, we learn patchwork, a ggplot2 extension designed to combine multiple ggplot2 plots into one figure.

> Patchwork has a simple syntax:

> - `+` arranges plots side by side 

> - `()` groups plots 

> - `/` stacks plots vertically

### Combining Two ggplot2 Graphs

The tabset below shows two histograms combined using patchwork by using `+`. The syntax is simple and intuitive.

:::{.panel-tabset}

#### Left & right

```{r}
#| code-fold: true
#| code-summary: "VIEW CODE"
p1 + p2
```

#### Top & bottom

```{r}
#| code-fold: true
#| code-summary: "VIEW CODE"
p1 / p2
```
:::

### Combining Three ggplot2 Graphs

More complex composite plots can be created using patchwork operators. For example, the figure below uses `/` to stack plots, `|` to place plots side by side, and `()` to control the layout order.

Further examples can be found on [plot Assembly](https://patchwork.data-imaginist.com/articles/guides/assembly.html).

:::{.panel-tabset}
#### Three graphs

```{r, fig.width=12, fig.height=8}
#| code-fold: true
#| code-summary: "VIEW CODE"
(p1 / p2) | p3
```

#### Graph with table

Here, `wrap_table()` from **patchwork** is used to create a table. A similar result can also be achieved using `gt()` from the **gt** package.

```{r}
#| code-fold: true
#| code-summary: "VIEW CODE"
p3 + wrap_table(exam_data[1:10, c("MATHS", "ENGLISH")], panel = "full", space = "fixed") 
```
:::

### Creating a Composite Figure with Tag

To help reference subplots in the text, patchwork also supports auto-tagging, as shown in the figure below. Auto-tagging can be enabled by setting `tag_levels` in `plot_annotation()` to specify the tag style: `'1'` for Arabic numerals, `'A'` for uppercase letters, `'a'` for lowercase letters, `'I'` for uppercase Roman numerals, and `'i'` for lowercase Roman numerals.

:::{.panel-tabset}

#### Simple tagging

```{r, fig.width=12, fig.height=8}
#| code-fold: true
#| code-summary: "VIEW CODE"
((p1 / p2) | p3) +
  plot_annotation(
    tag_levels = "I",
    caption = "Disclaimer: These plots are mainly for practice.")
```

#### Tag with prefix and suffix

```{r, fig.width=12, fig.height=8}
#| code-fold: true
#| code-summary: "VIEW CODE"

((p1 / p2) | p3) + plot_annotation(tag_levels = c('A', '1'), tag_prefix = 'Fig. ',
                                 tag_sep = '.', tag_suffix = ':',caption = "Disclaimer: These plots are mainly for practice.")
```
:::

### Creating Figure with Insert

Besides arranging plots using predefined layouts, we could use [`inset_element()`](https://patchwork.data-imaginist.com/reference/inset_element.html) from **pacthwork** to place one or more plots or graphical elements freely on top of, or within, another plot.

::: callout-tip
#### Key arguments for inset_element()

-   `p`: object to inset (ggplot/grob/gt/etc.)

-   `left, bottom, right, top`: inset position (numeric or unit)

-   `align_to`: reference area ("panel" (default), "plot", "full")

-   `on_top`: place inset above or below the main plot

-   `clip`: clip inset to bounds or not

-   `ignore_tag`: exclude inset from auto-tagging
:::

:::{.panel-tabset}
#### Basic use

Tip: keep values between 0 and 1 (npc units) and always ensure left \< right, bottom \< top.

```{r}
#| code-fold: true
#| code-summary: "VIEW CODE"
#| 
p3 + inset_element(p2, 
                   left = 0.02, 
                   bottom = 0.7, 
                   right = 0.5, 
                   top = 1)
```

#### Alignment

```{r, fig.width=10, fig.height=8}
#| code-fold: true
#| code-summary: "VIEW CODE"

p3 + inset_element(p2, 
                   left = 0.5, 
                   bottom = 0.05, 
                   right = 1, 
                   top = 0.3,
                   align_to = "plot")
```

#### Tagging

```{r, fig.width=10, fig.height=8}
#| code-fold: true
#| code-summary: "VIEW CODE"

p3 + inset_element(p2, 0.5, 0.05, 1, 0.3, align_to= "plot",ignore_tag= TRUE) +
  plot_annotation(tag_levels = '1',  caption = "Demenstration only")
```
:::

### Creating a Composite Figure by Using patchwork and ggtheme

:::{.panel-tabset}
#### ggthemes

Figure below is created by combining patchwork and `theme_economist()` of ggthemes package using operator `&`, applying to all plots in the patchwork layout. If you use `+` , the theme is applied only to the last patchwork object, not every subplot.

```{r, fig.width=12, fig.height=8}
#| code-fold: true
#| code-summary: "VIEW CODE"
patchwork <- p3 / (p1 + p2) 
patchwork & theme_economist()
```

#### Built-in themes

```{r, fig.width=12, fig.height=8}
#| code-fold: true
#| code-summary: "VIEW CODE"
patchwork <- (p1 / p2) | p3
patchwork & theme_linedraw() 
```

#### hrbthemes

```{r, fig.width=12, fig.height=8}
#| code-fold: true
#| code-summary: "VIEW CODE"
patchwork <- (p1 / p2) | p3
patchwork  & theme_modern_rc(grid="Y")
```
:::

## Reference

Kam, Tin Seong (2026). [2 Beyond ggplot2 Fundamentals.](https://r4va.netlify.app/chap02) R for Visual Analytics.

[Patchwork R package goes nerd viral](https://www.littlemissdata.com/blog/patchwork)

[ggplot2 Theme Elements Demonstration](https://henrywang.nl/ggplot2-theme-elements-demonstration/)[s](https://albert-rapp.de/post/2021-10-28-extend-plot-variety/)

[ggplot2 Theme Elements Reference Sheet](https://isabella-b.com/blog/ggplot2-theme-elements-reference/)
