---
title: "05-4  Visual Multivariate Analysis with Parallel Coordinates Plot"
author: "Yiqiong PAN"
date: "09 Feb, 2026"
date-modified: last-modified

execute:
  echo: true    # display the code
  eval: true    # run the code
  message: false
  warning: false
  freeze: true  # do not re-render if nothing edited

format: 
  html:
    toc: true
    code-link: true

number-sections: true
number-depth: 6

editor: visual
---

## Overview

Parallel coordinates plots are designed for visualising and analysing multivariate numerical data. They allow multiple variables to be compared at once and help reveal relationships, such as factors contributing to a Happiness Index. The technique was introduced by Alfred Inselberg in the 1970s and is more common in academic and scientific settings than business reporting. As [Stephen Few](https://www.perceptualedge.com/articles/b-eye/parallel_coordinates.pdf) (2006) notes, parallel coordinates are most effective as an interactive exploration tool for uncovering multivariate patterns.

We will learn how to create static parallel coordinates plots using `ggparcoord()` from **GGally**, and interactive parallel coordinates plots using the **parallelPlot** package.

## Installing and Loading the Required Libraries

There are 3 packages that will be used in this exercise.

| Package | Description |
|----|----|
| [tidyverse](https://www.tidyverse.org/) | A collection of R packages for data importing, cleaning, transformation, and visualisation (e.g., [readr](https://readr.tidyverse.org/), dplyr, tidyr, [ggplot2](https://ggplot2.tidyverse.org/)). |
| [GGally](https://ggobi.github.io/ggally/index.html) | A ggplot2 extension for tidy and simplified multivariate visualisation |
| [parallelPlot](https://cran.r-project.org/web/packages/parallelPlot/parallelPlot.pdf) | Creates interactive parallel coordinates plots. |

The code below installs and loads the required packages into R environment.

```{r}
pacman::p_load(parallelPlot, GGally,tidyverse)
```

## Data Preparation

### The Data

In this hands on exercise, the [World Happiness 2018 Report](https://www.worldhappiness.report/ed/2018/) is used and downloaded from [here](https://s3.amazonaws.com/happiness-report/2018/WHR2018Chapter2OnlineData.xls). The original excel file is converted and saved as csv file and named `WHData-2018.csv`.

### Importing the Data

The code below uses `read_csv()` from **readr** package to import the dataset into R environment It contains 156 observations.

```{r}
wh <- read_csv("data/WHData-2018.csv")
```

The preview table displays the first six records of the dataset. It contains 12 columns, where **Country**, **Region** are categorical variables, while the remaining columns are numeric variables of type double that describe Happiness score, such as GDP per capita, Dystopia, Social support, and Healthy life expectancy.

```{r}
knitr::kable(head(wh, n=6),format = "html")
```

## Plotting Static Parallel Coordinates Plot

[`ggparcoord()`](https://ggobi.github.io/ggally/reference/ggparcoord.html) from **GGally** package is used to create static arallel coordinates plot.

::: callout-tips
### Key Arguments

-   `data`: the data.frame to plot.

-   `columns` : which variables (names or indices) become the parallel axes.

-   `groupColumn` : a variable (e.g. region) to colour (group) the lines by.

-   `scale` : how to scale variables (e.g., `"std"`, `"uniminmax"`(min 0, max 1), `"globalminmax"`, `"robust"`).

-   `scaleSummary` : summary statistic used when centring (`scale = "center"`).

-   `centerObsID` : observation to centre on when `scale = "centerObs"`.

-   `missing` : how to handle missing values (`"exclude"`, `"random"`, etc.).

-   `order` : how to order the axes (can be a custom order or methods like `"anyClass"`).

-   `showPoints` : if TRUE, adds points at each variable value.

-   `splineFactor` : whether to smooth lines using splines.

-   `alphaLines` : transparency (or a column name for per-line alpha) for the lines ( 0 -1).

-   `boxplot` / `shadeBox` : show boxplots or shaded ranges under each axis (TRUE or FALSE).

-   `mapping` : additional `ggplot2::aes()` mappings passed to the plot.

-   `title` : plot title.
:::

:::: panel-tabset
### Plotting a simple parallel coordinates plot

The code below creates a basic static parrallel coordinators plot based on columns 7 to 12.

```{r, fig.width=12}
ggparcoord(data = wh, columns = c(7:12))
```

### Plotting a parallel coordinates with boxplot

The basic parallel coordinates plot did not provide a meaningful interpretation of the World Happiness indicators. Next we improve the plot but adding a range of arguments available in `ggparcoord()`.

```{r}
wh$Region <- as.factor(wh$Region)
```

```{r, fig.width=12, fig.height=6}
ggparcoord(data = wh, columns = c(7:12),
           groupColumn = "Region",  # Use the column name instead of index
           scale = "uniminmax",
           alphaLines = 0.2,
           boxplot = TRUE,
           title = "Parallel Coordinates Plot of World Happiness Variables")
```

### Parallel coordinates with facet

Since `ggparcoord()` extends **ggplot2**, we can combine it with **ggplot2** functions when creating parallel coordinates plots.

In the code chunk below, `facet_wrap()` is used to produce 10 small-multiple parallel coordinates plots, with each panel representing one geographical region (for example, East Asia). One aesthetic issue with the current design is that some variable names overlap on the x-axis.

```{r, fig.width=12, fig.height=6}
ggparcoord(data = wh, 
           columns = c(7:12), 
           groupColumn = 2,
           scale = "uniminmax",
           alphaLines = 0.2,
           boxplot = TRUE, 
           title = "Multiple Parallel Coordinates Plots of World Happines Variables by Region") +
  facet_wrap(~ Region)
```

### Rotating x-axis text label

we use theme() function from ggplot2 to retate the axis text labels by 30 degrees to make it easy to read.

```{r, fig.width=14, fig.height=6}
ggparcoord(data = wh, 
           columns = c(7:12), 
           groupColumn = 2,
           scale = "uniminmax",
           alphaLines = 0.2,
           boxplot = TRUE, 
           title = "Multiple Parallel Coordinates Plots of World Happines Variables by Region") +
  facet_wrap(~ Region)+
  theme(axis.text.x = element_text(angle = 30, vjust = 0.8, size = 5))
```

::: callout-tips
#### Note

-   To rotate the x-axis text labels, use the `axis.text.x` argument in the `theme()` function, and specify `element_text(angle = 30)` to rotate the x-axis labels by 30 degrees. we can also contronl the y-axis using `axis.text.y`.

-   `hjust`(left 0, centre 0.5, right 1) and `vjust`(bottom 0, centre 0.5, top 1) control text alignment relative to its position, and they are especially useful after rotating axis labels.

-   We can also control the axis text `size` using `axis.text.x` and `axis.text.y` in `theme()`.
:::

### Adjusting the rotated x-axis text label

Rotating the x-axis text labels to 30 degrees causes them to overlap with the plot area. This can be avoided by adjusting the text position using the `hjust` argument in `element_text()` within the `theme()` function. Since we are modifying the appearance of the x-axis labels, we use `axis.text.x`.

```{r, fig.width=14, fig.height=6}
ggparcoord(data = wh, 
           columns = c(7:12), 
           groupColumn = 2,
           scale = "uniminmax",
           alphaLines = 0.2,
           boxplot = TRUE, 
           title = "Multiple Parallel Coordinates Plots of World Happines Variables by Region") +
  facet_wrap(~ Region)+
  theme(axis.text.x = element_text(angle = 30, hjust = 1, size = 5))
```
::::

## Plotting Interactive Parallel Coordinates Plot: parallelPlot Methods

[`parallelPlot`](https://cran.r-project.org/web/packages/parallelPlot/) is an R package designed specifically for creating parallel coordinates plots using the `htmlwidgets` framework and [d3.js](https://d3js.org/). In this section, we learn how to use the functions in `parallelPlot` to build interactive parallel coordinates plots.

::: callout-tips
### Key Arguments of parallelplot()

Here is the refined version with **arguments and options wrapped in backticks**, using **colons**, and no references.

Key arguments of `parallelPlot()`:

-   `data`: a data frame containing the variables to be visualised

-   `categorical`: a named list specifying which columns are categorical and their corresponding levels

-   `refColumnDim`: the column name used to colour the lines, which can be continuous or categorical

-   `continuousCS`: the colour scale applied when `refColumnDim` is continuous

-   `categoricalCS`: the colour scale applied when `refColumnDim` is categorical

-   `categoriesRep`: controls how categorical variables are displayed, such as `EquallySizedBoxes` or `EquallySpacedLines`

-   `arrangeMethod`: determines how lines are arranged within categories to minimise crossings, for example `fromLeft`, `fromRight`, `fromBoth`, or `fromNone`

-   `inputColumns`: specifies which columns are treated as input dimensions in the plot

-   `histoVisibility`: controls whether histograms are displayed on each axis

-   `invertedAxes`: specifies which axes should be inverted or flipped

-   `cutoffs`: defines threshold values for variables to filter or highlight observations

-   `refRowIndex`: specifies the row index used as a reference baseline line

-   `rotateTitle`: controls whether axis titles are rotated

-   `columnLabels`: allows custom axis labels, including HTML formatting

-   `sliderPosition`: controls which dimensions are initially visible when there are many axes

-   `controlWidgets`: enables interactive control widgets, commonly used with Shiny

\-`cssRules`: allows custom CSS styling of the parallel coordinates plot

-   `width` and `height`: specify the size of the interactive widget in pixels
:::

### The basic plot

The code below shows how to use `parallelPlot()` to create a simple interactive paprallel coordinates plot.

First we select the required columns.

```{r}
wh <- wh %>%
  select("Happiness score", c(7:12))

```

After executing the code, we notice that some of the axis labels overlap with each other.

```{r, fig.width=16, fig.height=14}
parallelPlot(wh, width = 320, height = 250)
```

### Rotate axis label

We enable `rotateTitle` argument to avoid overlapping axis labels.

```{r, fig.width=16, fig.height=14}
parallelPlot(wh, rotateTitle = TRUE)
```

One useful interactive feature of `parallelPlot` is that when we click on a variable of interest, for example the Happiness score, the default monotonous blue colour changes to a blue colour scale with varying intensity.

### Changing the colour scheme

Since the data is on a **continuous scale**, we can change the default blue colour by using the `continuousCS` argument to assign a specific colour scheme to the chart. If the data is **discrete (categorical)**, we can instead use the `categoricalCS` argument to apply an appropriate categorical colour scheme.

```{r, fig.width=16, fig.height=14}
parallelPlot(wh, rotateTitle = TRUE, continuousCS = "YlOrRd")
```

If we click on a continuous variable, we observe that the colour scheme changes from the default Viridis to a yellow–orange–red gradient, with the colours scaled by intensity.

### Parallel coordinates plot with histogram

The `histoVisibility` argument can be used to display histograms along the axis of each variable. `parallelPlot()` expects histoVisibility to be a vector, not a single logical value. The first line creates a logical vector where every element is `TRUE`, and its length equals the number of columns in `wh`.

```{r, fig.width=16, fig.height=14}
histoVisiblity <- rep(TRUE, ncol(wh))
parallelPlot(wh, rotateTitle = TRUE,
             histoVisibility = histoVisiblity)
```

### Enabling reference column dimension and flipped axes

Here we enable the reference dimension as Happiness score adn change the direction of the axis.

```{r, fig.width=14, fig.height=12}

parallelPlot(
  wh,
  rotateTitle = TRUE,
  refColumnDim = "Happiness score",
  invertedAxes = c("Freedom", "Corruption"))

```

### Filtering the data before creating the plot

Here we filter the observations with happiness score between 4 and 7.

```{r, fig.width=14, fig.height=12}
wh_filtered <- wh[wh$`Happiness score` >= 4 & wh$`Happiness score` <= 7, ]

parallelPlot(
  wh_filtered,
  rotateTitle = TRUE
)

```

## Reference

Kam, Tin Seong (2026). [15 Visual Multivariate Analysis with Parallel Coordinates Plot](https://r4va.netlify.app/chap15) R for Visual Analytics.
