---
title: "Extra_1: Time on the Horizon: ggHoriPlot methods"
author: "Yiqiong PAN"
date: "27 Feb, 2026"
date-modified: last-modified

execute:
  echo: true    # display the code
  eval: true    # run the code
  message: false
  warning: false
  freeze: true  # do not re-render if nothing edited

format: 
  html:
    toc: true
    code-link: true

number-sections: true
number-depth: 6

editor: visual
---

## Learning Outcome

A horizon graph is a multi-panel area chart that visualises a numeric data over a continuous interval such as a time period. It is useful for understanding the overall trend, patterns and outliers in the data.


## Getting Started

### Installing and Loading the Required Libraries

There are **4** packages that will be used in this exercise.

| Package | Description |
|------------------------------------|------------------------------------|
| [tidyverse](https://www.tidyverse.org/) | A collection of R packages for data importing, cleaning, transformation, and visualisation (e.g., [readr](https://readr.tidyverse.org/), dplyr, tidyr, [ggplot2](https://ggplot2.tidyverse.org/)). |
| [lubridate](https://cran.r-project.org/web/packages/lubridate/index.html) | A Tidyverse package that handles date-time objects |
| [ggthemes](https://cran.r-project.org/web/packages/ggthemes/index.html) | Provides 'ggplot2' themes and scales that replicate some of the famous looks. |
| [ggHoriPlot](https://rivasiker.github.io/ggHoriPlot/) | Extends horizon plots in ggplot2 |

The code below installs and loads the required packages into R environment.

```{r}
pacman::p_load(tidyverse, ggthemes,lubridate, ggHoriPlot)
```

## Step 1: Data Import


For this hands-on exercise, the [**Average Retail Prices of Selected Consumer Items**](https://tablebuilder.singstat.gov.sg/page-not-found) file will be used. It contains 7,452 rows of time series records of average retail prices for selected consumer items.

### Importing the Data

We use `read_csv()` from the **readr** package to import the dataset *AVERP.csv* into the R environment as a tibble named `df`. We then preview the dataset using `kable()`.

```{r}
averp <- read_csv("data/AVERP.csv")
```

The dataset contains three columns: two categorical variables, **Date** and **Consumer Items**, and one numeric column **Values**. Since we want to perform time series plotting, it is essential to convert the **Date** to proper date data type. Here we use [`dmy()`](https://lubridate.tidyverse.org/reference/ymd.html) from [**lubridate**](https://lubridate.tidyverse.org/) pacakge to convert the column into date format. 

```{r}
averp <- averp %>%
  mutate(`Date` = dmy(`Date`))
```


```{r}
class(averp$`Consumer Items`)
class(averp$Date)
```

```{r}
knitr::kable(head(averp), "html")
```


### Step 2. Plotting the horizon graph


**Horizon plots** compare each consumer item to its own history/pattern of average retail prices while **line plots** with same scales compare the average prices to each other. The code below creates a horizon plot by using [`geom_horizon()`](https://rivasiker.github.io/ggHoriPlot/reference/geom_horizon.html) from [**ggHoriPlot**](https://rivasiker.github.io/ggHoriPlot/index.html) package. "%b%y" in `date_labels` controls the format of date that displays under the x-axis. "3-month" of `date_breaks`set the labels only shows every 3 months.


```{r, fig.height=12, fig.width=10}
averp %>%
  filter(Date >= "2018-01-01") %>%
  ggplot() +
  geom_horizon(aes(x = Date, y = Values),
               origin = "midpoint",
               horizonscale = 6) +
facet_grid(`Consumer Items`~.) +
  theme_few() +
  scale_fill_hcl(palette = 'RdBu') +
  theme(panel.spacing.y = unit(0, "lines"),
        strip.text.y = element_text(
          size = 5, angle = 0, hjust = 0
        ),
        legend.position = 'none',
        axis.text.y = element_blank(),
        axis.text.x = element_text(size = 7),
        axis.title.y = element_blank(),
        axis.title.x = element_blank(),
        axis.ticks.y = element_blank(),
        axis.ticks.x = element_blank()) +
  scale_x_date(expand = c(0,0), date_breaks = "3 month", date_labels = "%b%y") +
  ggtitle('Average Retail Prices of Selected Consumer Items(Jan 2018 to Dec 2022)')
  
```

## Reference

Kam, Tin Seong (2026). 17 Visualising and Analysing Time-oriented Data, R for Visual Analytics, https://r4va.netlify.app/chap17
